---
title: "05_GSE181919_T Cell Exhaustion Analysis"
author: "Koushika R"
date: "2026-03-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
knitr::opts_knit$set(root.dir = "~/GSE181919_HNSCC")
```


# STEP 1: Define Paths
```{r paths}
base_dir   <- "~/GSE181919_HNSCC"
output_dir <- file.path(base_dir, "output")
plot_dir   <- file.path(base_dir, "plots", "Tcell_plots")
result_dir <- file.path(base_dir, "results")
study_id   <- "GSE181919"

dir.create(plot_dir,   showWarnings = FALSE, recursive = TRUE)
dir.create(result_dir, showWarnings = FALSE, recursive = TRUE)

cat("Paths defined ✅\n")
```


# STEP 2: Load Libraries
```{r libraries}
# UCell: scoring gene modules (exhaustion signatures) per cell
#        Better than AddModuleScore — uses rank-based approach, not affected by cell number or gene expression scale

#BiocManager::install("UCell")
#install.packages('ggpubr')
#install.packages('ggridges')

library(Seurat)
library(ggplot2)
library(dplyr)
library(tidyr)
library(scales)
library(RColorBrewer)
library(UCell)         # Gene module scoring
library(ggpubr)        # Statistical comparisons on plots
library(ggridges)      # Ridge plots for score distributions

cat("Libraries loaded ✅\n")
```


# STEP 3: Load Annotated Seurat Object & Subset T Cells
```{r load_seurat}
# Input: 05_GSE181919_celltypist_annotated.rds
# We subset to T cells only using author cell.type labels
# T cells: 17,533 cells — the largest population in the dataset


cat("Loading Seurat object...\n")
seurat_obj <- readRDS(
    file.path(output_dir, "05_GSE181919_celltypist_annotated.rds")
)
seurat_obj <- UpdateSeuratObject(seurat_obj)
cat("Full object loaded:", ncol(seurat_obj), "cells\n")

# Subset T cells only
tcells <- subset(seurat_obj, subset = cell.type == "T.cells")
cat("T cells subset:", ncol(tcells), "cells\n")
cat("Tissue type distribution:\n")
print(table(tcells$tissue.type))
cat("\nHPV status distribution:\n")
print(table(tcells$hpv))

# Free full object from memory
rm(seurat_obj)
gc()
cat("Full Seurat object removed from memory ✅\n")
```


# STEP 4: Re-normalize and Re-cluster T Cells
```{r recluster_tcells}
# Re-clustering within T cells only gives finer resolution —
# reveals CD4+, CD8+, Treg, exhausted, naive, memory subclusters
# that were merged in the full-dataset clustering


cat("Re-normalizing T cells...\n")

# Re-normalize within T cell subset
tcells <- NormalizeData(tcells)
tcells <- FindVariableFeatures(tcells, nfeatures = 2000)
tcells <- ScaleData(tcells)
tcells <- RunPCA(tcells, npcs = 30)

# Cluster T cells
tcells <- FindNeighbors(tcells, dims = 1:20)
tcells <- FindClusters(tcells, resolution = 0.5)
tcells <- RunUMAP(tcells, dims = 1:20)

cat("T cell re-clustering complete ✅\n")
cat("T cell clusters:", length(unique(tcells$seurat_clusters)), "\n")
```


# STEP 5: Visualize T Cell Subclusters
```{r tcell_umap_clusters}
# Plot 1: T cell subclusters
p1 <- DimPlot(tcells,
              reduction  = "umap",
              group.by   = "seurat_clusters",
              label      = TRUE,
              label.size = 4,
              pt.size    = 0.3) +
    labs(title = paste0(study_id, " | T Cell Subclusters")) +
    theme(plot.title = element_text(hjust = 0.5, face = "bold"))

ggsave(file.path(plot_dir, paste0(study_id, "_Tcell_subclusters.png")),
       p1, width = 9, height = 7, dpi = 150, bg = "white")
p1

# Plot 2: T cells colored by tissue type
p2 <- DimPlot(tcells,
              reduction = "umap",
              group.by  = "tissue.type",
              pt.size   = 0.3) +
    scale_color_manual(values = c("NL" = "#4DAF4A", "LP" = "#FF7F00",
                                   "CA" = "#E41A1C", "LN" = "#984EA3")) +
    labs(title = paste0(study_id, " | T Cells by Tissue Type")) +
    theme(plot.title = element_text(hjust = 0.5, face = "bold"))

ggsave(file.path(plot_dir, paste0(study_id, "_Tcell_tissue_type.png")),
       p2, width = 9, height = 7, dpi = 150, bg = "white")
p2

# Plot 3: T cells colored by HPV status
p3 <- DimPlot(tcells,
              reduction = "umap",
              group.by  = "hpv",
              pt.size   = 0.3) +
    scale_color_manual(values = c("HPV+" = "#00BFC4", "HPV-" = "#F8766D")) +
    labs(title = paste0(study_id, " | T Cells by HPV Status")) +
    theme(plot.title = element_text(hjust = 0.5, face = "bold"))

ggsave(file.path(plot_dir, paste0(study_id, "_Tcell_HPV_status.png")),
       p3, width = 9, height = 7, dpi = 150, bg = "white")
p3
```


# STEP 6: Key T Cell Marker & Exhaustion Gene Feature Plots
```{r exhaustion_markers_featureplot}

# Canonical T cell state markers:
#   CD8A/CD4:      CD8+ cytotoxic vs CD4+ helper
#   FOXP3/IL2RA:   Regulatory T cells (Tregs)
#   PDCD1 (PD-1):  Exhaustion checkpoint — upregulated in exhausted T cells
#   HAVCR2 (TIM-3):Co-inhibitory receptor — marks terminal exhaustion
#   LAG3:          Inhibitory receptor — exhaustion marker
#   TIGIT:         Checkpoint — co-expressed with PD-1 in exhaustion
#   CTLA4:         Checkpoint — Treg marker + early exhaustion
#   TCF7:          Transcription factor — marks progenitor/stem-like exhausted T cells
#   GZMB/PRF1:     Cytotoxic effector molecules — functional T cells
#   MKI67:         Proliferation marker
# =============================================================================

exhaustion_genes <- c(
    "CD8A", "CD4", "FOXP3", "IL2RA",
    "PDCD1", "HAVCR2", "LAG3", "TIGIT", "CTLA4",
    "TCF7", "GZMB", "PRF1", "MKI67"
)

# Keep only genes present in dataset
exhaustion_genes <- exhaustion_genes[exhaustion_genes %in% rownames(tcells)]
cat("Genes available:", paste(exhaustion_genes, collapse = ", "), "\n")

p4 <- FeaturePlot(tcells,
                  features   = exhaustion_genes,
                  ncol       = 4,
                  pt.size    = 0.1,
                  order      = TRUE,
                  cols       = c("lightgrey", "#E41A1C")) &
    theme(plot.title   = element_text(size = 10, face = "bold"),
          axis.title   = element_blank(),
          axis.text    = element_blank(),
          axis.ticks   = element_blank())

ggsave(file.path(plot_dir, paste0(study_id, "_Tcell_exhaustion_featureplot.png")),
       p4, width = 16, height = 12, dpi = 150, bg = "white")
p4
```


# STEP 7: Score T Cell States Using UCell
```{r exhaustion_scores}
# UCell assigns a per-cell score for each gene module (0–1 scale)
# Higher score = stronger expression of that gene program
#
# Gene signatures:
#   Exhausted:      Core exhaustion checkpoint receptors
#   Effector:       Cytotoxic molecules — functional killing capacity
#   Naive/Memory:   Early/resting T cell markers
#   Treg:           Regulatory T cell signature
#   Progenitor_Tex: Stem-like exhausted T cells (TCF7+ PD-1+)
#                   — important for immunotherapy response
# =============================================================================

signatures <- list(
    Exhausted      = c("PDCD1", "HAVCR2", "LAG3", "TIGIT", "CTLA4"),
    Effector       = c("GZMB", "PRF1", "IFNG", "TNF", "GZMA"),
    Naive_Memory   = c("TCF7", "CCR7", "SELL", "LEF1", "IL7R"),
    Treg           = c("FOXP3", "IL2RA", "IKZF2", "CTLA4", "TNFRSF18"),
    Progenitor_Tex = c("TCF7", "PDCD1", "SLAMF6", "CXCR5", "TOX")
)

# Score all signatures
tcells <- AddModuleScore_UCell(tcells, features = signatures)

cat("UCell scores added ✅\n")
cat("Score columns:\n")
print(grep("_UCell", colnames(tcells@meta.data), value = TRUE))
```


# STEP 8: Visualize Exhaustion Scores on UMAP
```{r exhaustion_umap_scores}
score_cols <- grep("_UCell", colnames(tcells@meta.data), value = TRUE)

p5 <- FeaturePlot(tcells,
                  features = score_cols,
                  ncol     = 3,
                  pt.size  = 0.1,
                  order    = TRUE,
                  cols     = c("lightgrey", "#2166AC")) &
    theme(plot.title = element_text(size = 10, face = "bold"),
          axis.title = element_blank(),
          axis.text  = element_blank(),
          axis.ticks = element_blank())

ggsave(file.path(plot_dir, paste0(study_id, "_Tcell_state_scores_umap.png")),
       p5, width = 15, height = 10, dpi = 150, bg = "white")
p5
```


# STEP 9: Exhaustion Score — HPV+ vs HPV– Violin Plot
```{r exhaustion_by_hpv_violin}
# Key comparison: Are T cells more exhausted in HPV+ or HPV– tumors?
# Filter to CA only — HPV annotation is only meaningful in primary cancer
# Hypothesis:
#   HPV+ tumors are T cell-rich (from TME analysis) — chronic antigen
#   exposure from HPV may drive more exhaustion in HPV+ T cells
#   OR HPV+ T cells remain more functional due to stronger TCR signaling
# =============================================================================

tcells_ca <- subset(tcells, subset = tissue.type == "CA")
cat("CA T cells:", ncol(tcells_ca), "\n")
cat("HPV distribution in CA T cells:\n")
print(table(tcells_ca$hpv))

# Reshape for plotting
score_df <- tcells_ca@meta.data %>%
    select(hpv, all_of(score_cols)) %>%
    pivot_longer(cols      = all_of(score_cols),
                 names_to  = "State",
                 values_to = "Score") %>%
    mutate(State = gsub("_UCell", "", State))

# Violin plot with statistical comparison
p6 <- ggplot(score_df, aes(x = hpv, y = Score, fill = hpv)) +
    geom_violin(trim  = FALSE, alpha = 0.8) +
    geom_boxplot(width = 0.1, fill = "white", outlier.size = 0.3) +
    facet_wrap(~State, scales = "free_y", ncol = 3) +
    scale_fill_manual(values = c("HPV+" = "#00BFC4", "HPV-" = "#F8766D")) +
    stat_compare_means(method   = "wilcox.test",
                       label    = "p.signif",
                       hide.ns  = FALSE,
                       size     = 4) +
    labs(title = paste0(study_id, " | T Cell State Scores: HPV+ vs HPV– (CA only)"),
         x     = "HPV Status",
         y     = "UCell Score",
         fill  = "HPV Status") +
    theme_minimal(base_size = 12) +
    theme(plot.title    = element_text(hjust = 0.5, face = "bold"),
          strip.text    = element_text(face = "bold"),
          legend.position = "bottom")

ggsave(file.path(plot_dir, paste0(study_id, "_Tcell_scores_HPV_violin.png")),
       p6, width = 14, height = 10, dpi = 300, bg = "white")
p6
```


# STEP 10: Exhaustion Score Across Tissue Types — Ridge Plot
```{r exhaustion_by_tissue_violin}
# Shows how T cell exhaustion evolves across disease progression
# Expected: exhaustion score increases NL → LP → CA → LN
# as T cells encounter chronic antigen stimulation in tumor


score_df_tissue <- tcells@meta.data %>%
    select(tissue.type, all_of(score_cols)) %>%
    pivot_longer(cols      = all_of(score_cols),
                 names_to  = "State",
                 values_to = "Score") %>%
    mutate(State       = gsub("_UCell", "", State),
           tissue.type = factor(tissue.type,
                                levels = c("NL", "LP", "CA", "LN")))

# Ridge plot
p7 <- ggplot(score_df_tissue,
             aes(x = Score, y = tissue.type, fill = tissue.type)) +
    geom_density_ridges(alpha = 0.8, scale = 1.2) +
    facet_wrap(~State, scales = "free_x", ncol = 3) +
    scale_fill_manual(values = c("NL" = "#4DAF4A", "LP" = "#FF7F00",
                                  "CA" = "#E41A1C", "LN" = "#984EA3")) +
    labs(title = paste0(study_id, " | T Cell State Scores Across Disease Progression"),
         x     = "UCell Score",
         y     = "Tissue Type",
         fill  = "Tissue Type") +
    theme_minimal(base_size = 12) +
    theme(plot.title     = element_text(hjust = 0.5, face = "bold"),
          strip.text     = element_text(face = "bold"),
          legend.position = "bottom")

ggsave(file.path(plot_dir, paste0(study_id, "_Tcell_scores_tissue_ridge.png")),
       p7, width = 14, height = 10, dpi = 300, bg = "white")
p7
```


# STEP 11: Classify T Cells into Discrete States
```{r tcell_state_classification}
# Each T cell is assigned to a state based on its highest UCell score
# States: Exhausted / Effector / Naive_Memory / Treg / Progenitor_Tex

# This converts continuous scores → discrete labels for proportion analysis

# Assign state based on highest score
tcells$tcell_state <- apply(
    tcells@meta.data[, score_cols], 1,
    function(x) {
        state <- names(which.max(x))
        gsub("_UCell", "", state)
    }
)

cat("T cell state distribution:\n")
print(table(tcells$tcell_state))

cat("\nT cell states by tissue type:\n")
print(table(tcells$tissue.type, tcells$tcell_state))
```


# STEP 12: T Cell States on UMAP
```{r tcell_state_umap}
state_colors <- c(
    "Exhausted"      = "#E41A1C",
    "Effector"       = "#FF7F00",
    "Naive_Memory"   = "#4DAF4A",
    "Treg"           = "#984EA3",
    "Progenitor_Tex" = "#377EB8"
)

p8 <- DimPlot(tcells,
              reduction  = "umap",
              group.by   = "tcell_state",
              pt.size    = 0.3,
              label      = TRUE,
              label.size = 4) +
    scale_color_manual(values = state_colors) +
    labs(title = paste0(study_id, " | T Cell States")) +
    theme(plot.title = element_text(hjust = 0.5, face = "bold"))

ggsave(file.path(plot_dir, paste0(study_id, "_Tcell_states_umap.png")),
       p8, width = 10, height = 7, dpi = 150, bg = "white")
p8
```


# STEP 13: T Cell State Proportions — HPV+ vs HPV– and by Tissue Type
```{r tcell_state_proportions}

# Proportions by HPV status (CA only)
prop_hpv_state <- tcells@meta.data %>%
    filter(tissue.type == "CA") %>%
    group_by(hpv, tcell_state) %>%
    summarise(count = n(), .groups = "drop") %>%
    group_by(hpv) %>%
    mutate(proportion = count / sum(count) * 100)

p9 <- ggplot(prop_hpv_state,
             aes(x = hpv, y = proportion, fill = tcell_state)) +
    geom_bar(stat = "identity", width = 0.6) +
    scale_fill_manual(values = state_colors) +
    scale_y_continuous(labels = percent_format(scale = 1)) +
    labs(title = paste0(study_id, " | T Cell State Proportions: HPV+ vs HPV– (CA only)"),
         x     = "HPV Status",
         y     = "Proportion (%)",
         fill  = "T Cell State") +
    theme_minimal(base_size = 14) +
    theme(plot.title = element_text(hjust = 0.5, face = "bold"))

ggsave(file.path(plot_dir, paste0(study_id, "_Tcell_state_HPV_proportions.png")),
       p9, width = 8, height = 7, dpi = 300, bg = "white")
p9

# Proportions by tissue type
prop_tissue_state <- tcells@meta.data %>%
    group_by(tissue.type, tcell_state) %>%
    summarise(count = n(), .groups = "drop") %>%
    group_by(tissue.type) %>%
    mutate(proportion = count / sum(count) * 100)

p10 <- ggplot(prop_tissue_state,
              aes(x    = factor(tissue.type, levels = c("NL","LP","CA","LN")),
                  y    = proportion,
                  fill = tcell_state)) +
    geom_bar(stat = "identity", width = 0.7) +
    scale_fill_manual(values = state_colors) +
    scale_y_continuous(labels = percent_format(scale = 1)) +
    labs(title = paste0(study_id, " | T Cell State Proportions Across Progression"),
         x     = "Tissue Type",
         y     = "Proportion (%)",
         fill  = "T Cell State") +
    theme_minimal(base_size = 14) +
    theme(plot.title = element_text(hjust = 0.5, face = "bold"))

ggsave(file.path(plot_dir, paste0(study_id, "_Tcell_state_tissue_proportions.png")),
       p10, width = 10, height = 7, dpi = 300, bg = "white")
p10
```


# STEP 14: Exhaustion Marker Dot Plot per T Cell State
```{r exhaustion_dotplot}

# Validates discrete state assignments by showing marker gene expression
# per state — should show:
#   Exhausted state:      High PDCD1, HAVCR2, LAG3, TIGIT
#   Effector state:       High GZMB, PRF1, IFNG
#   Naive/Memory state:   High TCF7, CCR7, SELL, IL7R
#   Treg state:           High FOXP3, IL2RA, CTLA4
#   Progenitor_Tex state: High TCF7 + PDCD1 (unique co-expression)
# =============================================================================

Idents(tcells) <- "tcell_state"

validation_genes <- c(
    # Exhaustion
    "PDCD1", "HAVCR2", "LAG3", "TIGIT",
    # Effector
    "GZMB", "PRF1", "IFNG", "GZMA",
    # Naive/Memory
    "TCF7", "CCR7", "SELL", "IL7R",
    # Treg
    "FOXP3", "IL2RA", "IKZF2",
    # Progenitor
    "TOX", "SLAMF6", "CXCR5"
)

validation_genes <- validation_genes[validation_genes %in% rownames(tcells)]

p11 <- DotPlot(tcells,
               features  = validation_genes,
               cols      = c("lightgrey", "#2166AC"),
               dot.scale = 6,
               group.by  = "tcell_state") +
    coord_flip() +
    labs(title = paste0(study_id, " | Exhaustion Markers per T Cell State"),
         x     = "Gene",
         y     = "T Cell State") +
    theme_minimal(base_size = 11) +
    theme(plot.title  = element_text(hjust = 0.5, face = "bold"),
          axis.text.x = element_text(angle = 45, hjust = 1))

ggsave(file.path(plot_dir, paste0(study_id, "_Tcell_exhaustion_dotplot.png")),
       p11, width = 12, height = 10, dpi = 300, bg = "white")
p11
```


# STEP 15: Save Results
```{r save_results}

# Save T cell state proportions by tissue
write.csv(prop_tissue_state,
          file.path(result_dir, paste0(study_id,
                    "_Tcell_state_proportions_by_tissue.csv")),
          row.names = FALSE)

# Save T cell state proportions by HPV
write.csv(prop_hpv_state,
          file.path(result_dir, paste0(study_id,
                    "_Tcell_state_proportions_by_HPV.csv")),
          row.names = FALSE)

# Save T cell metadata with states and scores
write.csv(
    tcells@meta.data[, c("tissue.type", "hpv", "tcell_state", score_cols)],
    file.path(result_dir, paste0(study_id, "_Tcell_scores_and_states.csv"))
)

# Save T cell Seurat object
saveRDS(tcells,
        file.path(output_dir, "06_GSE181919_Tcells_exhaustion.rds"),
        version = 2)

cat("All results saved ✅\n")
cat("\n=== T Cell Exhaustion Summary ===\n")
cat("Total T cells analysed: ", ncol(tcells), "\n")
cat("T cell subclusters:     ", length(unique(tcells$seurat_clusters)), "\n")
cat("T cell states assigned: ", length(unique(tcells$tcell_state)), "\n")
cat("\nState distribution:\n")
print(table(tcells$tcell_state))
```

