---
title: "06_GSE181919_Differential Expression Analysis"
author: "Koushika R"
date: "2026-03-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
knitr::opts_knit$set(root.dir = "C:/Users/koush/Desktop/Koush/Projects/Workshops/Single Cell RNASeq HandsOn Workshop/GSE181919_HNSCC")
```


# STEP 1: Define Paths
```{r paths}
base_dir   <- "C:/Users/koush/Desktop/Koush/Projects/Workshops/Single Cell RNASeq HandsOn Workshop/GSE181919_HNSCC"
output_dir <- file.path(base_dir, "output")
plot_dir   <- file.path(base_dir, "plots", "DEG_plots")
result_dir <- file.path(base_dir, "results", "DEG")
study_id   <- "GSE181919"

dir.create(plot_dir,   showWarnings = FALSE, recursive = TRUE)
dir.create(result_dir, showWarnings = FALSE, recursive = TRUE)

cat("Paths defined ✅\n")
```


# STEP 2: Load Libraries
```{r libraries}
# DESeq2:     Pseudobulk DEG — gold standard for condition comparisons
#             Aggregates cells per sample → treats samples as replicates
#             Avoids false positives from treating cells as independent
# ggrepel:    Non-overlapping labels on volcano plots
# pheatmap:   Heatmap of top DEGs across conditions
# =============================================================================

#BiocManager::install('DESeq2')
#install.packages('ggrepel')
#install.packages('pheatmap')

library(Seurat)
library(ggplot2)
library(dplyr)
library(tidyr)
library(DESeq2)
library(ggrepel)
library(pheatmap)
library(RColorBrewer)
library(scales)

cat("Libraries loaded ✅\n")
```


# STEP 3: Load Annotated Seurat Object
```{r load_seurat}
# Input: 05_GSE181919_celltypist_annotated.rds
# Contains: cell.type (10 types), tissue.type (NL/LP/CA/LN),
#           hpv, patient.id, celltypist_label
# =============================================================================

cat("Loading Seurat object...\n")
seurat_obj <- readRDS(
    file.path(output_dir, "05_GSE181919_celltypist_annotated.rds")
)
seurat_obj <- UpdateSeuratObject(seurat_obj)

cat("Loaded:", ncol(seurat_obj), "cells ×", nrow(seurat_obj), "genes\n")
cat("Cell types:", paste(unique(seurat_obj$cell.type), collapse = ", "), "\n")
cat("Tissue types:", paste(unique(seurat_obj$tissue.type), collapse = ", "), "\n")
```

# STEP 4: Find Marker Genes per Cell Type (FindAllMarkers)
```{r marker_genes}
# Identifies genes that distinguish each cell type from all others
# Method: Wilcoxon rank-sum test (fast, non-parametric)
# Filters:
#   min.pct = 0.25    → gene expressed in ≥25% of cells in either group
#   logfc.threshold = 0.25 → minimum log2 fold change
# This is cell identity markers — NOT condition comparison

Idents(seurat_obj) <- "cell.type"

cat("Finding marker genes per cell type...\n")
#This may take 5-10 minutes for 51,849 cells

all_markers <- FindAllMarkers(
    seurat_obj,
    only.pos       = TRUE,
    min.pct        = 0.25,
    logfc.threshold = 0.25,
    test.use       = "wilcox"
)

cat("Done ✅\n")
cat("Total marker genes found:", nrow(all_markers), "\n")

# Top 10 markers per cell type
top10_markers <- all_markers %>%
    group_by(cluster) %>%
    slice_max(order_by = avg_log2FC, n = 10) %>%
    ungroup()

cat("\nTop 10 markers per cell type:\n")
print(top10_markers %>% select(cluster, gene, avg_log2FC, p_val_adj))

# Save all markers
write.csv(all_markers,
          file.path(result_dir, paste0(study_id, "_all_cell_type_markers.csv")),
          row.names = FALSE)
cat("\nMarkers saved ✅\n")
```


# STEP 5: Heatmap of Top Marker Genes per Cell Type
```{r marker_heatmap}
# Visualizes top 5 marker genes per cell type
# Color = scaled average expression per cell type

top5_markers <- all_markers %>%
    group_by(cluster) %>%
    slice_max(order_by = avg_log2FC, n = 5) %>%
    ungroup()

# Downsample for plotting speed
set.seed(88)
seurat_down <- subset(seurat_obj,
                      cells = sample(colnames(seurat_obj), 5000))

p1 <- DoHeatmap(seurat_down,
                features = top5_markers$gene,
                group.by = "cell.type",
                size     = 3,
                angle    = 45) +
    scale_fill_gradientn(colors = c("#2166AC", "white", "#E41A1C")) +
    labs(title = paste0(study_id, " | Top 5 Marker Genes per Cell Type")) +
    theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 12),
          axis.text.y = element_text(size = 6))

ggsave(file.path(plot_dir, paste0(study_id, "_DEG_marker_heatmap.png")),
       p1, width = 16, height = 12, dpi = 150, bg = "white")
p1
```


```{r Quick_check}
# Quick diagnostic — check what AggregateExpression actually names its columns

# Test with T cells only
test_sub <- subset(seurat_obj,
                   cells = which(seurat_obj$cell.type == "T.cells" &
                                 seurat_obj$tissue.type %in% c("NL","CA")))

# Build sample_id
test_sub$sample_id <- paste0(test_sub$patient.id, "_",
                              test_sub$tissue.type)

cat("Sample IDs we built:\n")
print(head(unique(test_sub$sample_id), 10))

# Run AggregateExpression
test_counts <- AggregateExpression(
    test_sub,
    group.by      = "sample_id",
    assays        = "RNA",
    slot          = "counts",
    return.seurat = FALSE
)$RNA

cat("\nColumn names from AggregateExpression:\n")
print(head(colnames(test_counts), 10))

cat("\nDo they match?\n")
cat("Built sample_ids:", length(unique(test_sub$sample_id)), "\n")
cat("AggregateExpression cols:", ncol(test_counts), "\n")
```



# STEP 6: Pseudobulk Setup for Condition DEG
```{r pseudobulk_setup}
# Why pseudobulk?
#   - Single-cell DEG methods (Wilcoxon) treat each cell as independent
#   - But cells from the same patient are NOT independent — they share patient-level variation (genetics, treatment history)
#   - Pseudobulk aggregates all cells of the same type per patient → sample
#   - This gives TRUE biological replicates (23 patients)
#   - DESeq2 then finds genes that differ between conditions across patients
#
# Comparisons planned:
#   1. NL vs CA  — normal vs primary cancer (key disease comparison)
#   2. CA vs LN  — primary cancer vs metastasis
#   3. HPV+ vs HPV– in CA T cells — key immunotherapy question
#
# We run DEG for major cell types with enough cells across conditions:
# Fix: AggregateExpression converts "_" → "-" in column names
# Solution: build sample_id with "-" to match output directly

make_pseudobulk <- function(seurat, celltype, group_col, groups,
                             patient_col  = "patient.id",
                             celltype_col = "cell.type") {

    keep <- which(seurat[[celltype_col]][,1] == celltype &
                  seurat[[group_col]][,1] %in% groups)
    sub  <- subset(seurat, cells = keep)
    cat("  Cells in", celltype, ":", ncol(sub), "\n")

    # Use "-" not "_" — AggregateExpression replaces "_" with "-"
    sub$sample_id <- paste0(
        sub[[patient_col]][,1], "-",
        sub[[group_col]][,1]
    )

    # Aggregate counts
    counts <- AggregateExpression(
        sub,
        group.by      = "sample_id",
        assays        = "RNA",
        slot          = "counts",
        return.seurat = FALSE
    )$RNA

    # Build colData from column names directly
    col_df <- data.frame(
        sample_id = colnames(counts),
        condition = sapply(strsplit(colnames(counts), "-"), tail, 1),
        row.names = colnames(counts)
    )

    cat("  Samples:", nrow(col_df),
        "| Conditions:", paste(unique(col_df$condition), collapse = "/"), "\n")

    list(counts = counts, coldata = col_df)
}

cat("Fixed pseudobulk function v2 defined ✅\n")
```


# STEP 7: DEG — NL vs CA per Cell Type (Pseudobulk DESeq2)
```{r deg_nl_vs_ca}
# Identifies genes that change between normal tissue and primary cancer per cell type — reveals cell-type-specific transcriptional reprogramming
# Reference = NL, Test = CA → positive LFC = upregulated in CA
# =============================================================================

# Cell types with sufficient cells in both NL and CA
celltypes_nlca <- c("T.cells", "Macrophages", "Fibroblasts",
                    "B_Plasma.cells", "Endothelial.cells")

deg_nlca_list <- list()

for (ct in celltypes_nlca) {
    cat("\nProcessing:", ct, "(NL vs CA)\n")

    tryCatch({
        pb     <- make_pseudobulk(seurat_obj, ct, "tissue.type", c("NL","CA"))
        counts <- pb$counts
        coldat <- pb$coldata

        # Keep only samples with ≥10 cells (already aggregated — use all)
        dds <- DESeqDataSetFromMatrix(
            countData = round(counts),
            colData   = coldat,
            design    = ~ condition
        )

        # Filter low-count genes
        dds <- dds[rowSums(counts(dds)) >= 10, ]
        dds$condition <- relevel(factor(dds$condition), ref = "NL")
        dds <- DESeq(dds, quiet = TRUE)

        res <- results(dds, contrast = c("condition", "CA", "NL"),
                       alpha = 0.05)
        res_df <- as.data.frame(res) %>%
            tibble::rownames_to_column("gene") %>%
            filter(!is.na(padj)) %>%
            arrange(padj)

        deg_nlca_list[[ct]] <- res_df
        cat("  Significant DEGs (padj<0.05):",
            sum(res_df$padj < 0.05, na.rm = TRUE), "\n")

        # Save per cell type
        write.csv(res_df,
                  file.path(result_dir,
                            paste0(study_id, "_DEG_NLvsCA_",
                                   gsub("\\.", "_", ct), ".csv")),
                  row.names = FALSE)

    }, error = function(e) cat("  Skipped:", ct, "-", e$message, "\n"))
}

cat("\nNL vs CA DEG complete ✅\n")
```


# STEP 8: DEG — CA vs LN per Cell Type (Pseudobulk DESeq2)
```{r deg_ca_vs_ln}
# Identifies genes that change during metastasis
# Reference = CA, Test = LN → positive LFC = upregulated in metastasis

celltypes_caln <- c("T.cells", "Macrophages", "B_Plasma.cells")

deg_caln_list <- list()

for (ct in celltypes_caln) {
    cat("\nProcessing:", ct, "(CA vs LN)\n")

    tryCatch({
        pb     <- make_pseudobulk(seurat_obj, ct, "tissue.type", c("CA","LN"))
        counts <- pb$counts
        coldat <- pb$coldata

        dds <- DESeqDataSetFromMatrix(
            countData = round(counts),
            colData   = coldat,
            design    = ~ condition
        )
        dds <- dds[rowSums(counts(dds)) >= 10, ]
        dds$condition <- relevel(factor(dds$condition), ref = "CA")
        dds <- DESeq(dds, quiet = TRUE)

        res <- results(dds, contrast = c("condition", "LN", "CA"),
                       alpha = 0.05)
        res_df <- as.data.frame(res) %>%
            tibble::rownames_to_column("gene") %>%
            filter(!is.na(padj)) %>%
            arrange(padj)

        deg_caln_list[[ct]] <- res_df
        cat("  Significant DEGs (padj<0.05):",
            sum(res_df$padj < 0.05, na.rm = TRUE), "\n")

        write.csv(res_df,
                  file.path(result_dir,
                            paste0(study_id, "_DEG_CAvsLN_",
                                   gsub("\\.", "_", ct), ".csv")),
                  row.names = FALSE)

    }, error = function(e) cat("  Skipped:", ct, "-", e$message, "\n"))
}

cat("\nCA vs LN DEG complete ✅\n")
```

# STEP 9: HPV+ vs HPV–
```{r deg_hpv}
# Fix: "HPV-" contains "-" which is our delimiter — use a lookup table
# instead of parsing condition back from column names


celltypes_hpv <- c("T.cells", "Macrophages")
deg_hpv_list  <- list()

seurat_ca <- subset(seurat_obj, subset = tissue.type == "CA")

make_pseudobulk_hpv <- function(seurat, celltype,
                                 patient_col  = "patient.id",
                                 celltype_col = "cell.type") {

    keep <- which(seurat[[celltype_col]][,1] == celltype &
                  seurat$hpv %in% c("HPV+", "HPV-"))
    sub  <- subset(seurat, cells = keep)
    cat("  Cells in", celltype, ":", ncol(sub), "\n")

    # Use patient-hpv as sample id — replace "+" and "-" with words
    sub$hpv_clean  <- ifelse(sub$hpv == "HPV+", "HPVpos", "HPVneg")
    sub$sample_id  <- paste0(sub[[patient_col]][,1], "-", sub$hpv_clean)

    # Store lookup: sample_id → hpv_clean BEFORE aggregating
    lookup <- sub@meta.data %>%
        select(sample_id, condition = hpv_clean) %>%
        distinct(sample_id, .keep_all = TRUE)

    counts <- AggregateExpression(
        sub,
        group.by      = "sample_id",
        assays        = "RNA",
        slot          = "counts",
        return.seurat = FALSE
    )$RNA

    # Build colData from lookup
    col_df <- data.frame(
        sample_id = colnames(counts),
        row.names = colnames(counts)
    ) %>%
        left_join(lookup, by = "sample_id") %>%
        as.data.frame()
    rownames(col_df) <- col_df$sample_id

    cat("  Samples:", nrow(col_df),
        "| Conditions:", paste(unique(col_df$condition), collapse = "/"), "\n")

    list(counts = counts, coldata = col_df)
}

for (ct in celltypes_hpv) {
    cat("\nProcessing:", ct, "(HPV+ vs HPV–)\n")

    tryCatch({
        pb     <- make_pseudobulk_hpv(seurat_ca, ct)
        counts <- pb$counts
        coldat <- pb$coldata

        dds <- DESeqDataSetFromMatrix(
            countData = round(counts),
            colData   = coldat,
            design    = ~ condition
        )
        dds <- dds[rowSums(counts(dds)) >= 10, ]

        # Reference = HPVneg, Test = HPVpos
        dds$condition <- relevel(factor(dds$condition), ref = "HPVneg")
        dds <- DESeq(dds, quiet = TRUE)

        res <- results(dds, contrast = c("condition", "HPVpos", "HPVneg"),
                       alpha = 0.05)
        res_df <- as.data.frame(res) %>%
            tibble::rownames_to_column("gene") %>%
            filter(!is.na(padj)) %>%
            arrange(padj)

        deg_hpv_list[[ct]] <- res_df
        cat("  Significant DEGs (padj<0.05):",
            sum(res_df$padj < 0.05, na.rm = TRUE), "\n")

        write.csv(res_df,
                  file.path(result_dir,
                            paste0(study_id, "_DEG_HPVpos_vs_HPVneg_",
                                   gsub("\\.", "_", ct), ".csv")),
                  row.names = FALSE)

    }, error = function(e) cat("  Skipped:", ct, "-", e$message, "\n"))
}

cat("\nHPV DEG complete ✅\n")
```


# STEP 10: Volcano Plots — NL vs CA per Cell Type
```{r volcano_nlca}
# X-axis: log2 fold change (CA vs NL)
# Y-axis: -log10(adjusted p-value)
# Red:    Upregulated in CA (log2FC > 1, padj < 0.05)
# Blue:   Downregulated in CA (log2FC < -1, padj < 0.05)
# Top 10 significant genes labeled per plot
# =============================================================================

make_volcano <- function(res_df, title,
                         lfc_thresh = 1, padj_thresh = 0.05,
                         n_label = 10) {
    res_df <- res_df %>%
        mutate(
            direction = case_when(
                log2FoldChange >  lfc_thresh & padj < padj_thresh ~ "Up",
                log2FoldChange < -lfc_thresh & padj < padj_thresh ~ "Down",
                TRUE ~ "NS"
            ),
            neg_log10_padj = -log10(padj)
        )

    # Top genes to label
    top_genes <- res_df %>%
        filter(direction != "NS") %>%
        slice_min(order_by = padj, n = n_label)

    ggplot(res_df, aes(x = log2FoldChange,
                       y = neg_log10_padj,
                       color = direction)) +
        geom_point(alpha = 0.6, size = 1) +
        geom_text_repel(data  = top_genes,
                        aes(label = gene),
                        size  = 3,
                        max.overlaps = 20) +
        scale_color_manual(values = c("Up"   = "#E41A1C",
                                       "Down" = "#377EB8",
                                       "NS"   = "grey70")) +
        geom_vline(xintercept = c(-lfc_thresh, lfc_thresh),
                   linetype = "dashed", color = "grey40") +
        geom_hline(yintercept = -log10(padj_thresh),
                   linetype = "dashed", color = "grey40") +
        labs(title    = title,
             x        = "log2 Fold Change",
             y        = "-log10(adjusted p-value)",
             color    = "Direction") +
        theme_minimal(base_size = 12) +
        theme(plot.title = element_text(hjust = 0.5, face = "bold"))
}

# Generate volcano for each cell type
for (ct in names(deg_nlca_list)) {
    p <- make_volcano(
        deg_nlca_list[[ct]],
        title = paste0(study_id, " | NL vs CA — ", ct)
    )
    ggsave(file.path(plot_dir,
                     paste0(study_id, "_volcano_NLvsCA_",
                            gsub("\\.", "_", ct), ".png")),
           p, width = 9, height = 7, dpi = 150, bg = "white")
    print(p)
}
```


# STEP 11: Volcano Plots — HPV+ vs HPV– (T cells & Macrophages)
```{r volcano_hpv}
# Most important comparison — molecular basis of HPV status differences
# Positive LFC = upregulated in HPV+


for (ct in names(deg_hpv_list)) {
    p <- make_volcano(
        deg_hpv_list[[ct]],
        title = paste0(study_id, " | HPV+ vs HPV– — ", ct, " (CA only)")
    )
    ggsave(file.path(plot_dir,
                     paste0(study_id, "_volcano_HPV_",
                            gsub("\\.", "_", ct), ".png")),
           p, width = 9, height = 7, dpi = 150, bg = "white")
    print(p)
}
```


# STEP 12: Heatmap of Top DEGs — NL vs CA in T Cells
```{r top_deg_heatmap}
# Shows expression of top 30 DEGs (15 up + 15 down in CA) across all tissue types — validates progression pattern

if ("T.cells" %in% names(deg_nlca_list)) {
    res_tcell <- deg_nlca_list[["T.cells"]]

    # Top 15 up and 15 down in CA
    top_up   <- res_tcell %>%
        filter(log2FoldChange > 1, padj < 0.05) %>%
        slice_min(padj, n = 15) %>% pull(gene)

    top_down <- res_tcell %>%
        filter(log2FoldChange < -1, padj < 0.05) %>%
        slice_min(padj, n = 15) %>% pull(gene)

    top_genes <- c(top_up, top_down)
    top_genes <- top_genes[top_genes %in% rownames(seurat_obj)]

    if (length(top_genes) > 0) {
        # Pseudobulk average expression per tissue type
        Idents(seurat_obj) <- "cell.type"
        tcell_sub <- subset(seurat_obj, idents = "T.cells")
        Idents(tcell_sub) <- "tissue.type"

        avg_exp <- AverageExpression(tcell_sub,
                                     features   = top_genes,
                                     assays     = "RNA",
                                     slot       = "data")$RNA
        avg_exp <- avg_exp[, c("NL","LP","CA","LN")]

        # Scale rows
        scaled_exp <- t(scale(t(avg_exp)))

        # Annotation for up/down
        gene_anno <- data.frame(
            Direction = ifelse(rownames(scaled_exp) %in% top_up,
                               "Up in CA", "Down in CA"),
            row.names = rownames(scaled_exp)
        )

        anno_colors <- list(
            Direction = c("Up in CA"   = "#E41A1C",
                          "Down in CA" = "#377EB8")
        )

        png(file.path(plot_dir,
                      paste0(study_id, "_DEG_heatmap_Tcell_NLvsCA.png")),
            width = 8, height = 10, units = "in", res = 150)
        pheatmap(scaled_exp,
                 annotation_row  = gene_anno,
                 annotation_colors = anno_colors,
                 color           = colorRampPalette(
                     c("#2166AC", "white", "#E41A1C"))(100),
                 cluster_cols    = FALSE,
                 fontsize_row    = 8,
                 fontsize_col    = 12,
                 main            = paste0(study_id,
                                          " | Top DEGs NL vs CA — T Cells"))
        dev.off()
        cat("T cell DEG heatmap saved ✅\n")
    }
}
```


# STEP 13: DEG Summary Table
```{r deg_summary}
summary_rows <- list()

for (ct in names(deg_nlca_list)) {
    df  <- deg_nlca_list[[ct]]
    sig <- df %>% filter(padj < 0.05)
    summary_rows[[length(summary_rows)+1]] <- data.frame(
        Comparison = "NL vs CA",
        Cell_Type  = ct,
        Total_Tested = nrow(df),
        Sig_DEGs   = nrow(sig),
        Up_in_CA   = sum(sig$log2FoldChange > 0),
        Down_in_CA = sum(sig$log2FoldChange < 0)
    )
}

for (ct in names(deg_caln_list)) {
    df  <- deg_caln_list[[ct]]
    sig <- df %>% filter(padj < 0.05)
    summary_rows[[length(summary_rows)+1]] <- data.frame(
        Comparison = "CA vs LN",
        Cell_Type  = ct,
        Total_Tested = nrow(df),
        Sig_DEGs   = nrow(sig),
        Up_in_CA   = sum(sig$log2FoldChange < 0),  # Up in LN = neg LFC
        Down_in_CA = sum(sig$log2FoldChange > 0)
    )
}

for (ct in names(deg_hpv_list)) {
    df  <- deg_hpv_list[[ct]]
    sig <- df %>% filter(padj < 0.05)
    summary_rows[[length(summary_rows)+1]] <- data.frame(
        Comparison = "HPV+ vs HPV–",
        Cell_Type  = ct,
        Total_Tested = nrow(df),
        Sig_DEGs   = nrow(sig),
        Up_in_CA   = sum(sig$log2FoldChange > 0),  # Up in HPV+
        Down_in_CA = sum(sig$log2FoldChange < 0)   # Up in HPV–
    )
}

summary_df <- bind_rows(summary_rows)
names(summary_df)[5:6] <- c("Up_in_Test", "Up_in_Ref")

cat("=== DEG Summary ===\n")
print(summary_df)

write.csv(summary_df,
          file.path(result_dir,
                    paste0(study_id, "_DEG_summary.csv")),
          row.names = FALSE)
cat("Summary saved ✅\n")
```














