---
title: "07_GSE181919_CellChat_Cell-Cell_Communication_Analysis"
author: "Koushika R"
date: "2026-03-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
knitr::opts_knit$set(root.dir = "C:/Users/koush/Desktop/Koush/Projects/Workshops/Single Cell RNASeq HandsOn Workshop/GSE181919_HNSCC")
```


# STEP 1: Define Paths
```{r paths}
base_dir   <- "C:/Users/koush/Desktop/Koush/Projects/Workshops/Single Cell RNASeq HandsOn Workshop/GSE181919_HNSCC"
output_dir <- file.path(base_dir, "output")
plot_dir   <- file.path(base_dir, "plots", "CellChat_plots")
result_dir <- file.path(base_dir, "results", "CellChat")
study_id   <- "GSE181919"

dir.create(plot_dir,   showWarnings = FALSE, recursive = TRUE)
dir.create(result_dir, showWarnings = FALSE, recursive = TRUE)

cat("Paths defined ✅\n")
```


# STEP 2: Load Libraries
```{r libraries}
# CellChat: ligand-receptor interaction analysis
#   - Uses curated database of ~2,000 ligand-receptor pairs
#   - Computes communication probability per L-R pair per cell type pair
#   - Aggregates into signaling pathways
#   - Allows comparison across conditions
# =============================================================================

BiocManager::install("ComplexHeatmap")
library(ComplexHeatmap)
remotes::install_github('jinworks/CellChat')
library(CellChat)
install.packages("devtools")
devtools::install_github("immunogenomics/presto")
library(presto)

library(Seurat)
library(ggplot2)
library(dplyr)
library(patchwork)
library(RColorBrewer)

options(stringsAsFactors = FALSE)

cat("Libraries loaded ✅\n")
```


# STEP 3: Load Annotated Seurat Object
```{r load_seurat}
cat("Loading Seurat object...\n")
seurat_obj <- readRDS(
    file.path(output_dir, "05_GSE181919_celltypist_annotated.rds")
)
seurat_obj <- UpdateSeuratObject(seurat_obj)

cat("Loaded:", ncol(seurat_obj), "cells\n")
cat("Cell types:", paste(sort(unique(seurat_obj$cell.type)), collapse = ", "), "\n")
cat("Tissue types:", paste(unique(seurat_obj$tissue.type), collapse = ", "), "\n")
```


# STEP 4: Define Cell Type Colors
```{r color_setup}
# Consistent colors for all CellChat plots

celltype_colors <- c(
    "T.cells"            = "#E41A1C",
    "B_Plasma.cells"     = "#FF7F00",
    "Macrophages"        = "#984EA3",
    "Dendritic.cells"    = "#999999",
    "Mast.cells"         = "#00CED1",
    "Fibroblasts"        = "#377EB8",
    "Endothelial.cells"  = "#F781BF",
    "Malignant.cells"    = "#A65628",
    "Epithelial.cells"   = "#4DAF4A",
    "Myocytes"           = "#FFFF33"
)

cat("Colors defined ✅\n")
```


# STEP 5: CellChat Helper Function
```{r cellchat_function}
# Creates and processes a CellChat object from a Seurat subset
# Pipeline:
#   1. Extract normalized counts + metadata
#   2. Create CellChat object
#   3. Set ligand-receptor database (human, Secreted Signaling + ECM + Cell-Cell)
#   4. Pre-process expression data
#   5. Compute communication probabilities
#   6. Filter weak interactions
#   7. Compute pathway-level aggregated signals
#   8. Calculate network centrality (which cells are hubs)
# =============================================================================

run_cellchat <- function(seurat_sub, group_col = "cell.type",
                          min_cells = 10) {

    # Remove cell types with too few cells
    ct_counts <- table(seurat_sub[[group_col]][,1])
    keep_cts  <- names(ct_counts[ct_counts >= min_cells])
    seurat_sub <- subset(seurat_sub,
                         cells = which(seurat_sub[[group_col]][,1]
                                       %in% keep_cts))
    cat("  Cell types retained:", paste(sort(keep_cts), collapse=", "), "\n")
    cat("  Total cells:", ncol(seurat_sub), "\n")

    # Create CellChat object
    data_input <- GetAssayData(seurat_sub, assay = "RNA", layer = "data")
    meta       <- seurat_sub@meta.data[, group_col, drop = FALSE]
    colnames(meta) <- "labels"

    cc <- createCellChat(object   = data_input,
                         meta     = meta,
                         group.by = "labels")

    # Set human ligand-receptor database
    CellChatDB       <- CellChatDB.human
    CellChatDB.use   <- subsetDB(CellChatDB,
                                 search  = c("Secreted Signaling",
                                             "ECM-Receptor",
                                             "Cell-Cell Contact"),
                                 key     = "annotation")
    cc@DB <- CellChatDB.use

    # Pre-process
    cc <- subsetData(cc)
    cc <- identifyOverExpressedGenes(cc)
    cc <- identifyOverExpressedInteractions(cc)

    # Compute communication probabilities
    cc <- computeCommunProb(cc, type = "triMean")
    cc <- filterCommunication(cc, min.cells = min_cells)

    # Pathway-level signals
    cc <- computeCommunProbPathway(cc)
    cc <- aggregateNet(cc)

    # Network centrality
    cc <- netAnalysis_computeCentrality(cc, slot.name = "netP")

    cat("  CellChat object complete ✅\n")
    return(cc)
}

cat("CellChat helper function defined ✅\n")
```


#=======Steps 6 to 9 takes ~5 to 8 mins each to run completely===================

# STEP 6: Run CellChat — NL (Normal Tissue)
```{r cellchat_nl}
cat("Processing NL...\n")
seurat_nl <- subset(seurat_obj, subset = tissue.type == "NL")
cc_nl     <- run_cellchat(seurat_nl)

saveRDS(cc_nl,
        file.path(result_dir, paste0(study_id, "_cellchat_NL.rds")))
cat("NL CellChat saved ✅\n")
```


# STEP 7: Run CellChat — CA (Primary Cancer)
```{r cellchat_ca}
cat("Processing CA...\n")
seurat_ca <- subset(seurat_obj, subset = tissue.type == "CA")
cc_ca     <- run_cellchat(seurat_ca)

saveRDS(cc_ca,
        file.path(result_dir, paste0(study_id, "_cellchat_CA.rds")))
cat("CA CellChat saved ✅\n")
```


# STEP 8: Run CellChat — LN (Lymph Node Metastasis)
```{r cellchat_ln}
cat("Processing LN...\n")
seurat_ln <- subset(seurat_obj, subset = tissue.type == "LN")
cc_ln     <- run_cellchat(seurat_ln)

saveRDS(cc_ln,
        file.path(result_dir, paste0(study_id, "_cellchat_LN.rds")))
cat("LN CellChat saved ✅\n")
```


# STEP 9: Run CellChat — HPV+ and HPV– (CA only)
```{r cellchat_hpv}
cat("Processing HPV+ (CA only)...\n")
seurat_hpvpos <- subset(seurat_obj,
                         subset = tissue.type == "CA" & hpv == "HPV+")
cc_hpvpos     <- run_cellchat(seurat_hpvpos)
saveRDS(cc_hpvpos,
        file.path(result_dir, paste0(study_id, "_cellchat_HPVpos.rds")))
cat("HPV+ CellChat saved ✅\n")

cat("Processing HPV– (CA only)...\n")
seurat_hpvneg <- subset(seurat_obj,
                         subset = tissue.type == "CA" & hpv == "HPV-")
cc_hpvneg     <- run_cellchat(seurat_hpvneg)
saveRDS(cc_hpvneg,
        file.path(result_dir, paste0(study_id, "_cellchat_HPVneg.rds")))
cat("HPV– CellChat saved ✅\n")
```


# STEP 10: Interaction Summary — NL vs CA vs LN
```{r summary_numbers}
# Number of interactions and total interaction strength per condition
cat("=== Interaction Summary ===\n")
for (nm in c("NL","CA","LN")) {
    cc  <- get(paste0("cc_", tolower(nm)))
    cat(nm, "— Interactions:", cc@net$count %>% sum(),
        "| Strength:", round(cc@net$weight %>% sum(), 2), "\n")
}

# Comparison object for NL vs CA
object_list_nlca <- list(NL = cc_nl, CA = cc_ca)
cc_merged_nlca   <- mergeCellChat(object_list_nlca,
                                   add.names = c("NL","CA"))

# Comparison object for CA vs LN
object_list_caln <- list(CA = cc_ca, LN = cc_ln)
cc_merged_caln   <- mergeCellChat(object_list_caln,
                                   add.names = c("CA","LN"))

# Comparison object for HPV
object_list_hpv  <- list(HPVneg = cc_hpvneg, HPVpos = cc_hpvpos)
cc_merged_hpv    <- mergeCellChat(object_list_hpv,
                                   add.names = c("HPVneg","HPVpos"))

cat("\nMerged CellChat objects created ✅\n")
```


# STEP 11: Compare Total Interactions — NL vs CA vs LN
```{r interaction_counts_plot}
# Bar plots showing how total number and strength of interactions
# changes across disease progression

png(file.path(plot_dir,
              paste0(study_id, "_CellChat_interaction_counts_NLvsCA.png")),
    width = 12, height = 5, units = "in", res = 150)
compareInteractions(cc_merged_nlca,
                    show.legend  = FALSE,
                    group        = c(1,2),
                    color.use    = c("#4DAF4A","#E41A1C"))
dev.off()

png(file.path(plot_dir,
              paste0(study_id, "_CellChat_interaction_counts_CAvsLN.png")),
    width = 12, height = 5, units = "in", res = 150)
compareInteractions(cc_merged_caln,
                    show.legend  = FALSE,
                    group        = c(1,2),
                    color.use    = c("#E41A1C","#984EA3"))
dev.off()

cat("Interaction count plots saved ✅\n")
```


# STEP 12: Circle Plots — NL vs CA Communication Networks
```{r circle_plots_nlca}
# Each node = cell type, each edge = communication
# Edge width = interaction strength
# Arrows show direction: sender → receiver


# Number of interactions
png(file.path(plot_dir,
              paste0(study_id, "_CellChat_circle_NL_count.png")),
    width = 8, height = 8, units = "in", res = 150)
netVisual_circle(cc_nl@net$count,
                 vertex.weight  = table(cc_nl@idents),
                 weight.scale   = TRUE,
                 label.edge     = FALSE,
                 title.name     = "NL — Number of Interactions",
                 color.use      = celltype_colors[names(cc_nl@idents %>%
                                    levels() %>% intersect(names(celltype_colors)))])
dev.off()

png(file.path(plot_dir,
              paste0(study_id, "_CellChat_circle_CA_count.png")),
    width = 8, height = 8, units = "in", res = 150)
netVisual_circle(cc_ca@net$count,
                 vertex.weight  = table(cc_ca@idents),
                 weight.scale   = TRUE,
                 label.edge     = FALSE,
                 title.name     = "CA — Number of Interactions")
dev.off()

cat("Circle plots saved ✅\n")
```

# STEP 13: Heatmap — Differential Interactions
```{r heatmap_nlca}
# Fix: netVisual_heatmap requires both objects to have identical cell types
# Solution: use measure = "count" with the raw net slot directly,
# OR use netVisual_heatmap with the merged object's shared cell types
# =============================================================================

# Check cell types in each
cat("NL cell types:", paste(levels(cc_nl@idents), collapse=", "), "\n")
cat("CA cell types:", paste(levels(cc_ca@idents), collapse=", "), "\n")
cat("LN cell types:", paste(levels(cc_ln@idents), collapse=", "), "\n")

# Find shared cell types across all three
shared_nlca <- intersect(levels(cc_nl@idents), levels(cc_ca@idents))
shared_caln <- intersect(levels(cc_ca@idents), levels(cc_ln@idents))

cat("\nShared NL/CA:", paste(shared_nlca, collapse=", "), "\n")
cat("Shared CA/LN:", paste(shared_caln, collapse=", "), "\n")

# Subset each CellChat object to shared cell types before merging
cc_nl_sub <- subsetCellChat(cc_nl, idents.use = shared_nlca)
cc_ca_sub <- subsetCellChat(cc_ca, idents.use = shared_nlca)

cc_ca_sub2 <- subsetCellChat(cc_ca, idents.use = shared_caln)
cc_ln_sub  <- subsetCellChat(cc_ln, idents.use = shared_caln)

# Re-merge with matched cell types
cc_merged_nlca <- mergeCellChat(list(NL = cc_nl_sub, CA = cc_ca_sub),
                                 add.names = c("NL","CA"))
cc_merged_caln <- mergeCellChat(list(CA = cc_ca_sub2, LN = cc_ln_sub),
                                 add.names = c("CA","LN"))

cat("\nRe-merged with shared cell types ✅\n")

# Now plot
png(file.path(plot_dir,
              paste0(study_id, "_CellChat_heatmap_diff_NLvsCA.png")),
    width = 10, height = 8, units = "in", res = 150)
netVisual_heatmap(cc_merged_nlca,
                  color.heatmap = c("#377EB8","white","#E41A1C"),
                  title.name    = "NL vs CA — Differential Interactions")
dev.off()

png(file.path(plot_dir,
              paste0(study_id, "_CellChat_heatmap_diff_CAvsLN.png")),
    width = 10, height = 8, units = "in", res = 150)
netVisual_heatmap(cc_merged_caln,
                  color.heatmap = c("#377EB8","white","#E41A1C"),
                  title.name    = "CA vs LN — Differential Interactions")
dev.off()

cat("Differential heatmaps saved ✅\n")
```



# STEP 14: Top Signaling Pathways in CA
```{r signaling_pathways_ca}
# Dot plot of top pathways ranked by communication probability
# Dot size = number of ligand-receptor pairs
# Color = total communication probability
# =============================================================================

# Get top pathways in CA
top_pathways_ca <- cc_ca@netP$pathways
cat("Top pathways in CA (", length(top_pathways_ca), "total):\n")
print(head(top_pathways_ca, 20))

# Focus on exhaustion + stromal pathways
exhaustion_pathways <- c("PD-L1", "MHC-II", "MHC-I", "CXCL",
                          "TIGIT", "PDCD1", "CD80", "CTLA4",
                          "GALECTIN", "CD226", "CD96")
stromal_pathways    <- c("TGFb", "VEGF", "COLLAGEN", "FN1",
                          "LAMININ", "MMP", "SPP1", "PERIOSTIN",
                          "THBS", "ANGPT", "NOTCH")

focus_pathways <- c(exhaustion_pathways, stromal_pathways)
found_pathways <- focus_pathways[focus_pathways %in% top_pathways_ca]

cat("\nFocus pathways found in CA:\n")
print(found_pathways)
```

# STEP 15 (FIXED): Bubble Plots — CA only
```{r bubble_plots_ca}

# Bubble plot 1: Exhaustion pathways → T cells
if (length(exhaustion_found) > 0) {
    png(file.path(plot_dir,
                  paste0(study_id, "_CellChat_bubble_exhaustion_CA.png")),
        width = 16, height = 10, units = "in", res = 200, bg = "white")
    p <- netVisual_bubble(
        cc_ca,
        targets.use    = "T.cells",
        signaling      = exhaustion_found,
        remove.isolate = TRUE,
        angle.x        = 45,
        title.name     = "CA — Exhaustion Pathways → T Cells"
    )
    print(p)
    dev.off()
    cat("Exhaustion bubble saved ✅\n")
}

# Bubble plot 2: Stromal pathways from Fibroblasts
if (length(stromal_found) > 0) {
    png(file.path(plot_dir,
                  paste0(study_id, "_CellChat_bubble_stromal_CA.png")),
        width = 16, height = 10, units = "in", res = 200, bg = "white")
    p <- netVisual_bubble(
        cc_ca,
        sources.use    = "Fibroblasts",
        signaling      = stromal_found,
        remove.isolate = TRUE,
        angle.x        = 45,
        title.name     = "CA — Stromal Pathways from Fibroblasts"
    )
    print(p)
    dev.off()
    cat("Stromal bubble saved ✅\n")
}
```


# STEP 16: Network Centrality — Sender/Receiver/Mediator/Influencer
```{r centrality_plots}
# Identifies which cell types are communication HUBS in NL vs CA
# Sender:    primary signal source
# Receiver:  primary signal target
# Mediator:  relays signals between others
# Influencer: modulates network-wide communication
# =============================================================================

png(file.path(plot_dir,
              paste0(study_id, "_CellChat_centrality_NL.png")),
    width = 10, height = 6, units = "in", res = 150)
netAnalysis_signalingRole_scatter(cc_nl,
    title = "NL — Signaling Roles")
dev.off()

png(file.path(plot_dir,
              paste0(study_id, "_CellChat_centrality_CA.png")),
    width = 10, height = 6, units = "in", res = 150)
netAnalysis_signalingRole_scatter(cc_ca,
    title = "CA — Signaling Roles")
dev.off()

png(file.path(plot_dir,
              paste0(study_id, "_CellChat_centrality_LN.png")),
    width = 10, height = 6, units = "in", res = 150)
netAnalysis_signalingRole_scatter(cc_ln,
    title = "LN — Signaling Roles")
dev.off()

cat("Centrality plots saved ✅\n")
```


# STEP 17 (FIXED): HPV+ vs HPV–  Differential Communication + Bubble
```{r hpv_comparison}
cat("=== HPV Interaction Summary ===\n")
cat("HPV– interactions:", cc_hpvneg@net$count %>% sum(), "\n")
cat("HPV+ interactions:", cc_hpvpos@net$count %>% sum(), "\n")

# Differential heatmap
png(file.path(plot_dir,
              paste0(study_id, "_CellChat_heatmap_diff_HPV.png")),
    width = 10, height = 8, units = "in", res = 150, bg = "white")
netVisual_heatmap(cc_merged_hpv,
                  color.heatmap = c("#F8766D","white","#00BFC4"),
                  title.name    = "HPV– vs HPV+ — Differential Interactions")
dev.off()
cat("HPV heatmap saved ✅\n")

# HPV bubble plot — exhaustion pathways → T cells
# Find pathways present in BOTH HPV objects
exhaustion_hpv_found <- intersect(
    exhaustion_found,
    intersect(cc_hpvpos@netP$pathways, cc_hpvneg@netP$pathways)
)
cat("Exhaustion pathways in both HPV groups:",
    paste(exhaustion_hpv_found, collapse=", "), "\n")

if (length(exhaustion_hpv_found) > 0) {
    png(file.path(plot_dir,
                  paste0(study_id, "_CellChat_bubble_exhaustion_HPV.png")),
        width = 16, height = 10, units = "in", res = 200, bg = "white")
    p <- netVisual_bubble(
        cc_merged_hpv,
        targets.use    = "T.cells",
        signaling      = exhaustion_hpv_found,
        comparison     = c(1, 2),
        remove.isolate = TRUE,
        angle.x        = 45,
        title.name     = "HPVneg vs HPVpos — Exhaustion Pathways → T Cells"
    )
    print(p)
    dev.off()
    cat("HPV exhaustion bubble saved ✅\n")
}
```


# STEP 18: Signaling Role Heatmap — All Pathways in CA
```{r pathway_heatmap}
# Fix: limit to top 20 pathways, larger canvas, bigger font


# Get top 20 pathways by total outgoing strength
top20 <- cc_ca@netP$pathways[1:min(20, length(cc_ca@netP$pathways))]
cat("Top 20 pathways:\n")
print(top20)

# Outgoing (sender) heatmap
png(file.path(plot_dir,
              paste0(study_id, "_CellChat_pathway_heatmap_CA_sender.png")),
    width = 14, height = 10, units = "in", res = 200, bg = "white")
netAnalysis_signalingRole_heatmap(
    cc_ca,
    pattern       = "outgoing",
    signaling     = top20,
    title         = "CA — Outgoing Signaling per Cell Type (Top 20)",
    color.heatmap = "OrRd",
    font.size     = 11,
    font.size.title = 14,
    width         = 14,
    height        = 10
)
dev.off()
cat("Sender heatmap saved ✅\n")

# Incoming (receiver) heatmap
png(file.path(plot_dir,
              paste0(study_id, "_CellChat_pathway_heatmap_CA_receiver.png")),
    width = 14, height = 10, units = "in", res = 200, bg = "white")
netAnalysis_signalingRole_heatmap(
    cc_ca,
    pattern       = "incoming",
    signaling     = top20,
    title         = "CA — Incoming Signaling per Cell Type (Top 20)",
    color.heatmap = "GnBu",
    font.size     = 11,
    font.size.title = 14,
    width         = 14,
    height        = 10
)
dev.off()
cat("Receiver heatmap saved ✅\n")
```


# STEP 19: Save Pathway Summaries as CSV
```{r save_results}
# Save pathway lists per condition
for (nm in c("NL","CA","LN","HPVpos","HPVneg")) {
    cc  <- get(paste0("cc_", tolower(nm)))
    df  <- data.frame(
        condition = nm,
        pathway   = cc@netP$pathways,
        stringsAsFactors = FALSE
    )
    write.csv(df,
              file.path(result_dir,
                        paste0(study_id, "_CellChat_pathways_", nm, ".csv")),
              row.names = FALSE)
}

# Save interaction summary
summary_df <- data.frame(
    Condition   = c("NL","CA","LN","HPVpos","HPVneg"),
    N_Cells     = c(ncol(seurat_nl), ncol(seurat_ca), ncol(seurat_ln),
                    ncol(seurat_hpvpos), ncol(seurat_hpvneg)),
    N_Interactions = c(
        sum(cc_nl@net$count),    sum(cc_ca@net$count),
        sum(cc_ln@net$count),    sum(cc_hpvpos@net$count),
        sum(cc_hpvneg@net$count)
    ),
    Total_Strength = c(
        round(sum(cc_nl@net$weight),2),    round(sum(cc_ca@net$weight),2),
        round(sum(cc_ln@net$weight),2),    round(sum(cc_hpvpos@net$weight),2),
        round(sum(cc_hpvneg@net$weight),2)
    ),
    N_Pathways  = c(
        length(cc_nl@netP$pathways),    length(cc_ca@netP$pathways),
        length(cc_ln@netP$pathways),    length(cc_hpvpos@netP$pathways),
        length(cc_hpvneg@netP$pathways)
    )
)

cat("=== CellChat Summary ===\n")
print(summary_df)

write.csv(summary_df,
          file.path(result_dir,
                    paste0(study_id, "_CellChat_summary.csv")),
          row.names = FALSE)
cat("\nAll results saved ✅\n")
```

