---
title: "08_GSE181919_Trajectory_Analysis_(Monocle3)"
author: "Koushika R"
date: "2026-03-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
knitr::opts_knit$set(root.dir = "~/GSE181919_HNSCC")
```


# STEP 1: Define Paths
```{r paths}
base_dir   <- "~/GSE181919_HNSCC"
output_dir <- file.path(base_dir, "output")
plot_dir   <- file.path(base_dir, "plots", "Trajectory_plots")
result_dir <- file.path(base_dir, "results", "Trajectory")
study_id   <- "GSE181919"

dir.create(plot_dir,   showWarnings = FALSE, recursive = TRUE)
dir.create(result_dir, showWarnings = FALSE, recursive = TRUE)

cat("Paths defined ✅\n")
```


# STEP 2: Libraries
```{r libraries}
# Monocle3: pseudotime trajectory inference
#   - Learns a principal graph through cell state space
#   - Assigns each cell a pseudotime value (0 = root/start state)
#   - Root = manually defined based on biology (naive T cells)
#   - Does NOT require tissue type ordering — learns order from transcriptomics
#   - SeuratWrappers bridges Seurat ↔ Monocle3
# =============================================================================

#Install grr from archive (no longer on CRAN)
#install.packages(
#    "https://cran.r-project.org/src/contrib/Archive/grr/grr_0.9.5.tar.gz",
#    repos = NULL,
#    type  = "source"
#)

#Install Bioconductor dependencies
#BiocManager::install(c(
#    "BiocGenerics", "DelayedArray", "DelayedMatrixStats",
#    "limma", "lme4", "S4Vectors", "SingleCellExperiment",
#    "SummarizedExperiment", "batchelor", "HDF5Array", "terra"
#), update = FALSE, ask = FALSE)

#Install monocle3 from GitHub
#remotes::install_github(
#    "cole-trapnell-lab/monocle3",
#    force = TRUE
#)

#Install SeuratWrappers
#remotes::install_github(
#    "satijalab/seurat-wrappers",
#    force = TRUE
#)

# Verify
library(monocle3)
library(SeuratWrappers)
cat("monocle3 version:", as.character(packageVersion("monocle3")), "\n")
cat("SeuratWrappers version:", as.character(packageVersion("SeuratWrappers")), "\n")

library(Seurat)
library(ggplot2)
library(dplyr)
library(patchwork)
library(RColorBrewer)
library(viridis)
library(pheatmap)

cat("Libraries loaded ✅\n")
```


# STEP 3: Load Seurat Objects
```{r load_data}
cat("Loading full annotated Seurat object...\n")
seurat_obj <- readRDS(
    file.path(output_dir, "05_GSE181919_celltypist_annotated.rds")
)
seurat_obj <- UpdateSeuratObject(seurat_obj)

cat("Loading T cell subset with exhaustion scores...\n")
seurat_tcell <- readRDS(
    file.path(output_dir, "06_GSE181919_Tcells_exhaustion.rds")
)
seurat_tcell <- UpdateSeuratObject(seurat_tcell)

cat("Full object:", ncol(seurat_obj), "cells\n")
cat("T cell subset:", ncol(seurat_tcell), "cells\n")
cat("T cell states:", paste(sort(unique(seurat_tcell$tcell_state)), collapse=", "), "\n")
```


# STEP 4: Color Palettes
```{r color_setup}
# Cell type colors (consistent with Scripts 04-07)
celltype_colors <- c(
    "T.cells"            = "#E41A1C",
    "B_Plasma.cells"     = "#FF7F00",
    "Macrophages"        = "#984EA3",
    "Dendritic.cells"    = "#999999",
    "Mast.cells"         = "#00CED1",
    "Fibroblasts"        = "#377EB8",
    "Endothelial.cells"  = "#F781BF",
    "Malignant.cells"    = "#A65628",
    "Epithelial.cells"   = "#4DAF4A",
    "Myocytes"           = "#FFFF33"
)

# Tissue type colors
tissue_colors <- c(
    "NL" = "#2CA02C",
    "LP" = "#FFBB78",
    "CA" = "#D62728",
    "LN" = "#9467BD"
)

# T cell state colors
state_colors <- c(
    "Exhausted"       = "#D62728",
    "Effector"        = "#FF7F0E",
    "Naive_Memory"    = "#2CA02C",
    "Treg"            = "#9467BD",
    "Progenitor_Tex"  = "#1F77B4"
)

# HPV colors
hpv_colors <- c("HPV+" = "#E41A1C", "HPV-" = "#377EB8")

cat("Colors defined ✅\n")
```


# STEP 5: Full TME Trajectory — All Cell Types
```{r full_tme_trajectory}
# Convert full Seurat object → Monocle3 CDS
# Use existing UMAP embedding from Harmony-corrected object
# Learn principal graph across all cell types
# This gives the macro-level view of TME evolution
# =============================================================================

cat("Converting full Seurat → Monocle3 CDS...\n")

# Use RNA assay for trajectory
DefaultAssay(seurat_obj) <- "RNA"

# Convert using SeuratWrappers
cds_full <- as.cell_data_set(seurat_obj)

# Transfer UMAP coordinates from Seurat
reducedDim(cds_full, "UMAP") <- Embeddings(seurat_obj, "umap")

# Cluster partitions (required by Monocle3)
# Use existing Seurat clusters as partitions
cds_full <- cluster_cells(cds_full, reduction_method = "UMAP",
                           resolution = 1e-3)

# Learn the principal graph
cat("Learning principal graph (full TME)...\n")
cds_full <- learn_graph(cds_full,
                         use_partition    = TRUE,
                         verbose          = FALSE,
                         learn_graph_control = list(
                             minimal_branch_len = 10,
                             ncenter           = 100
                         ))

cat("Principal graph learned ✅\n")
```


# STEP 6: Order Cells — Full TME
```{r order_cells_full}
# Root = cells in NL tissue (normal tissue = biological start)
# Monocle3 orders cells along the graph starting from root
# =============================================================================

# Define root: NL T cells as the biological starting point
# (normal tissue naive T cells are the most primitive state)
root_cells <- colnames(seurat_obj)[
    seurat_obj$tissue.type == "NL" &
    seurat_obj$cell.type   == "T.cells"
]
cat("Root cells (NL T cells):", length(root_cells), "\n")

# Order cells using defined root
cds_full <- order_cells(cds_full, root_cells = root_cells)

cat("Cells ordered by pseudotime ✅\n")
cat("Pseudotime range:",
    round(min(pseudotime(cds_full)[is.finite(pseudotime(cds_full))]), 2),
    "to",
    round(max(pseudotime(cds_full)[is.finite(pseudotime(cds_full))]), 2), "\n")
```


# STEP 7: Plot Full TME Trajectory
```{r plot_full_tme}
# Plot 1: Pseudotime on UMAP
png(file.path(plot_dir,
              paste0(study_id, "_trajectory_full_pseudotime.png")),
    width = 10, height = 8, units = "in", res = 200, bg = "white")
plot_cells(cds_full,
           color_cells_by          = "pseudotime",
           label_cell_groups       = FALSE,
           label_leaves            = FALSE,
           label_branch_points     = FALSE,
           graph_label_size        = 3,
           cell_size               = 0.4,
           trajectory_graph_color  = "black",
           trajectory_graph_segment_size = 0.8) +
    scale_color_viridis(option = "C", name = "Pseudotime") +
    ggtitle(paste0(study_id, " | Full TME Trajectory — Pseudotime")) +
    theme_classic(base_size = 12)
dev.off()

# Plot 2: Cell types on trajectory UMAP
png(file.path(plot_dir,
              paste0(study_id, "_trajectory_full_celltype.png")),
    width = 10, height = 8, units = "in", res = 200, bg = "white")
plot_cells(cds_full,
           color_cells_by          = "cell.type",
           label_cell_groups       = TRUE,
           label_leaves            = FALSE,
           label_branch_points     = FALSE,
           group_label_size        = 3.5,
           cell_size               = 0.4,
           trajectory_graph_color  = "grey50",
           trajectory_graph_segment_size = 0.6) +
    scale_color_manual(values = celltype_colors, name = "Cell Type") +
    ggtitle(paste0(study_id, " | Full TME Trajectory — Cell Types")) +
    theme_classic(base_size = 12)
dev.off()

# Plot 3: Tissue type on trajectory UMAP
png(file.path(plot_dir,
              paste0(study_id, "_trajectory_full_tissue.png")),
    width = 10, height = 8, units = "in", res = 200, bg = "white")
plot_cells(cds_full,
           color_cells_by          = "tissue.type",
           label_cell_groups       = FALSE,
           label_leaves            = FALSE,
           label_branch_points     = FALSE,
           cell_size               = 0.4,
           trajectory_graph_color  = "grey30",
           trajectory_graph_segment_size = 0.6) +
    scale_color_manual(values = tissue_colors, name = "Tissue") +
    ggtitle(paste0(study_id, " | Full TME Trajectory — Tissue Type")) +
    theme_classic(base_size = 12)
dev.off()

cat("Full TME trajectory plots saved ✅\n")
```


# STEP 8: Pseudotime Distribution by Tissue Type and Cell Type
```{r pseudotime_by_tissue}
# Violin plots showing whether cells from different tissue types
# occupy different pseudotime positions — validates NL→CA→LN ordering
# =============================================================================

# Extract pseudotime + metadata
pt_df <- data.frame(
    pseudotime  = pseudotime(cds_full),
    cell_type   = seurat_obj$cell.type,
    tissue_type = seurat_obj$tissue.type,
    hpv_status  = seurat_obj$hpv,
    stringsAsFactors = FALSE
) %>%
    filter(is.finite(pseudotime)) %>%
    mutate(tissue_type = factor(tissue_type, levels = c("NL","LP","CA","LN")))

# Violin: pseudotime by tissue type (all cells)
p1 <- ggplot(pt_df, aes(x = tissue_type, y = pseudotime,
                          fill = tissue_type)) +
    geom_violin(scale = "width", trim = TRUE, alpha = 0.8) +
    geom_boxplot(width = 0.1, fill = "white", outlier.size = 0.2) +
    scale_fill_manual(values = tissue_colors, guide = "none") +
    labs(title = paste0(study_id, " | Pseudotime by Tissue Type"),
         x = "Tissue Type", y = "Pseudotime") +
    theme_classic(base_size = 12) +
    theme(plot.title = element_text(hjust = 0.5))

png(file.path(plot_dir,
              paste0(study_id, "_trajectory_pseudotime_by_tissue.png")),
    width = 8, height = 6, units = "in", res = 200, bg = "white")
print(p1)
dev.off()

# Violin: pseudotime by cell type
p2 <- ggplot(pt_df, aes(x = reorder(cell_type, pseudotime, median),
                          y = pseudotime, fill = cell_type)) +
    geom_violin(scale = "width", trim = TRUE, alpha = 0.8) +
    geom_boxplot(width = 0.1, fill = "white", outlier.size = 0.2) +
    scale_fill_manual(values = celltype_colors, guide = "none") +
    labs(title = paste0(study_id, " | Pseudotime by Cell Type"),
         x = "Cell Type", y = "Pseudotime") +
    theme_classic(base_size = 12) +
    theme(axis.text.x  = element_text(angle = 45, hjust = 1),
          plot.title   = element_text(hjust = 0.5))

png(file.path(plot_dir,
              paste0(study_id, "_trajectory_pseudotime_by_celltype.png")),
    width = 10, height = 6, units = "in", res = 200, bg = "white")
print(p2)
dev.off()

cat("Pseudotime distribution plots saved ✅\n")

# Print median pseudotime per tissue
cat("\n=== Median Pseudotime by Tissue ===\n")
pt_df %>%
    group_by(tissue_type) %>%
    summarise(median_pt = round(median(pseudotime), 2),
              mean_pt   = round(mean(pseudotime),   2),
              n_cells   = n()) %>%
    print()
```


# STEP 9: T Cell Trajectory — Primary Analysis
```{r tcell_trajectory}
# Subset to T cells only for high-resolution exhaustion trajectory
# Root = Naive_Memory state (biologically most primitive T cell state)
# Expected trajectory: Naive_Memory → Progenitor_Tex → Exhausted
#                                   ↘ Effector
#                      Treg as separate branch
# =============================================================================

cat("Converting T cell Seurat → Monocle3 CDS...\n")
DefaultAssay(seurat_tcell) <- "RNA"

cds_tcell <- as.cell_data_set(seurat_tcell)

# Transfer T cell UMAP from Seurat
reducedDim(cds_tcell, "UMAP") <- Embeddings(seurat_tcell, "umap")

# Cluster for graph partitions
cds_tcell <- cluster_cells(cds_tcell, reduction_method = "UMAP",
                             resolution = 1e-3)

# Learn graph with tighter parameters for T cell resolution
cat("Learning T cell principal graph...\n")
cds_tcell <- learn_graph(cds_tcell,
                          use_partition    = FALSE,  # single partition for T cells
                          verbose          = FALSE,
                          learn_graph_control = list(
                              minimal_branch_len = 5,
                              ncenter           = 50
                          ))

# Root = Naive_Memory T cells
root_tcells <- colnames(seurat_tcell)[
    seurat_tcell$tcell_state == "Naive_Memory"
]
cat("Root cells (Naive_Memory):", length(root_tcells), "\n")

cds_tcell <- order_cells(cds_tcell, root_cells = root_tcells)

cat("T cell pseudotime range:",
    round(min(pseudotime(cds_tcell)[is.finite(pseudotime(cds_tcell))]), 2),
    "to",
    round(max(pseudotime(cds_tcell)[is.finite(pseudotime(cds_tcell))]), 2), "\n")
cat("T cell trajectory complete ✅\n")
```


# STEP 10: Plot T Cell Trajectory
```{r plot_tcell_trajectory}
# Plot 1: T cell pseudotime
png(file.path(plot_dir,
              paste0(study_id, "_trajectory_Tcell_pseudotime.png")),
    width = 10, height = 8, units = "in", res = 200, bg = "white")
plot_cells(cds_tcell,
           color_cells_by          = "pseudotime",
           label_cell_groups       = FALSE,
           label_leaves            = TRUE,
           label_branch_points     = TRUE,
           graph_label_size        = 4,
           cell_size               = 0.5,
           trajectory_graph_color  = "black",
           trajectory_graph_segment_size = 1) +
    scale_color_viridis(option = "C", name = "Pseudotime") +
    ggtitle(paste0(study_id, " | T Cell Exhaustion Trajectory — Pseudotime")) +
    theme_classic(base_size = 12)
dev.off()

# Plot 2: T cell states on trajectory
png(file.path(plot_dir,
              paste0(study_id, "_trajectory_Tcell_state.png")),
    width = 10, height = 8, units = "in", res = 200, bg = "white")
plot_cells(cds_tcell,
           color_cells_by          = "tcell_state",
           label_cell_groups       = TRUE,
           label_leaves            = TRUE,
           label_branch_points     = TRUE,
           group_label_size        = 4,
           cell_size               = 0.5,
           trajectory_graph_color  = "grey30",
           trajectory_graph_segment_size = 1) +
    scale_color_manual(values = state_colors, name = "T Cell State") +
    ggtitle(paste0(study_id, " | T Cell Exhaustion Trajectory — States")) +
    theme_classic(base_size = 12)
dev.off()

# Plot 3: Tissue type on T cell trajectory
png(file.path(plot_dir,
              paste0(study_id, "_trajectory_Tcell_tissue.png")),
    width = 10, height = 8, units = "in", res = 200, bg = "white")
plot_cells(cds_tcell,
           color_cells_by          = "tissue.type",
           label_cell_groups       = FALSE,
           label_leaves            = FALSE,
           label_branch_points     = FALSE,
           cell_size               = 0.5,
           trajectory_graph_color  = "grey30",
           trajectory_graph_segment_size = 1) +
    scale_color_manual(values = tissue_colors, name = "Tissue") +
    ggtitle(paste0(study_id, " | T Cell Trajectory — Tissue Type")) +
    theme_classic(base_size = 12)
dev.off()

# Plot 4: HPV status on T cell trajectory
png(file.path(plot_dir,
              paste0(study_id, "_trajectory_Tcell_HPV.png")),
    width = 10, height = 8, units = "in", res = 200, bg = "white")
plot_cells(cds_tcell,
           color_cells_by          = "hpv",
           label_cell_groups       = FALSE,
           label_leaves            = FALSE,
           label_branch_points     = FALSE,
           cell_size               = 0.5,
           trajectory_graph_color  = "grey30",
           trajectory_graph_segment_size = 1) +
    scale_color_manual(values = hpv_colors, name = "HPV Status") +
    ggtitle(paste0(study_id, " | T Cell Trajectory — HPV Status")) +
    theme_classic(base_size = 12)
dev.off()

cat("T cell trajectory plots saved ✅\n")
```


# STEP 11: T Cell Pseudotime Distributions
```{r tcell_pseudotime_dist}

pt_tcell <- data.frame(
    pseudotime  = pseudotime(cds_tcell),
    state       = seurat_tcell$tcell_state,
    tissue      = seurat_tcell$tissue.type,
    hpv         = seurat_tcell$hpv,
    stringsAsFactors = FALSE
) %>%
    filter(is.finite(pseudotime)) %>%
    mutate(tissue = factor(tissue, levels = c("NL","LP","CA","LN")),
           state  = factor(state,
                           levels = c("Naive_Memory","Progenitor_Tex",
                                      "Effector","Exhausted","Treg")))

# Print summary
cat("=== T Cell Pseudotime by State ===\n")
pt_tcell %>%
    group_by(state) %>%
    summarise(median_pt = round(median(pseudotime), 2),
              mean_pt   = round(mean(pseudotime),   2),
              n_cells   = n()) %>%
    arrange(median_pt) %>%
    print()

cat("\n=== T Cell Pseudotime by Tissue ===\n")
pt_tcell %>%
    group_by(tissue) %>%
    summarise(median_pt = round(median(pseudotime), 2),
              n_cells   = n()) %>%
    print()

# Plot: pseudotime by state (violin)
p_state <- ggplot(pt_tcell, aes(x = state, y = pseudotime, fill = state)) +
    geom_violin(scale = "width", trim = TRUE, alpha = 0.85) +
    geom_boxplot(width = 0.08, fill = "white", outlier.size = 0.3) +
    scale_fill_manual(values = state_colors, guide = "none") +
    labs(title = paste0(study_id, " | T Cell Pseudotime by Exhaustion State"),
         x = "T Cell State", y = "Pseudotime") +
    theme_classic(base_size = 12) +
    theme(axis.text.x = element_text(angle = 30, hjust = 1),
          plot.title  = element_text(hjust = 0.5))

png(file.path(plot_dir,
              paste0(study_id, "_trajectory_Tcell_pseudotime_by_state.png")),
    width = 9, height = 6, units = "in", res = 200, bg = "white")
print(p_state)
dev.off()

# Plot: pseudotime by tissue (violin)
p_tissue <- ggplot(pt_tcell, aes(x = tissue, y = pseudotime, fill = tissue)) +
    geom_violin(scale = "width", trim = TRUE, alpha = 0.85) +
    geom_boxplot(width = 0.08, fill = "white", outlier.size = 0.3) +
    scale_fill_manual(values = tissue_colors, guide = "none") +
    labs(title = paste0(study_id, " | T Cell Pseudotime by Tissue Type"),
         x = "Tissue", y = "Pseudotime") +
    theme_classic(base_size = 12) +
    theme(plot.title = element_text(hjust = 0.5))

png(file.path(plot_dir,
              paste0(study_id, "_trajectory_Tcell_pseudotime_by_tissue.png")),
    width = 8, height = 6, units = "in", res = 200, bg = "white")
print(p_tissue)
dev.off()

# Plot: pseudotime by HPV status (split violin)
p_hpv <- ggplot(pt_tcell %>% filter(tissue == "CA"),
                aes(x = hpv, y = pseudotime, fill = hpv)) +
    geom_violin(scale = "width", trim = TRUE, alpha = 0.85) +
    geom_boxplot(width = 0.08, fill = "white", outlier.size = 0.3) +
    scale_fill_manual(values = hpv_colors, guide = "none") +
    labs(title = paste0(study_id,
                        " | T Cell Pseudotime HPV+ vs HPV– (CA only)"),
         x = "HPV Status", y = "Pseudotime") +
    theme_classic(base_size = 12) +
    theme(plot.title = element_text(hjust = 0.5))

png(file.path(plot_dir,
              paste0(study_id, "_trajectory_Tcell_pseudotime_HPV.png")),
    width = 7, height = 6, units = "in", res = 200, bg = "white")
print(p_hpv)
dev.off()

cat("Pseudotime distribution plots saved ✅\n")
```


# STEP 12 : Key Exhaustion Gene Expression Along Pseudotime
```{r exhaustion_genes_on_trajectory}
# Plot canonical exhaustion genes on the T cell UMAP trajectory
# Shows WHERE on the trajectory each gene peaks
# =============================================================================

# Fix gene_short_name for both CDS objects (SeuratWrappers v0.4.0 quirk)
rowData(cds_tcell)$gene_short_name <- rownames(cds_tcell)
rowData(cds_full)$gene_short_name  <- rownames(cds_full)
cat("gene_short_name fixed for cds_tcell and cds_full ✅\n")

exhaustion_genes <- c(
    "CXCL13",   # Top DEG — exhausted T cell hallmark
    "PDCD1",    # PD-1 — canonical exhaustion marker
    "HAVCR2",   # TIM-3 — terminal exhaustion
    "LAG3",     # LAG-3 — exhaustion checkpoint
    "TIGIT",    # TIGIT — co-inhibitory receptor
    "TCF7",     # TCF1 — progenitor exhausted T cell marker
    "TOX",      # TOX — master exhaustion TF
    "FOXP3",    # FOXP3 — Treg marker
    "GZMB",     # Granzyme B — cytotoxic/effector
    "PRF1"      # Perforin — cytotoxic function
)

# Check which genes are present
available_genes <- exhaustion_genes[
    exhaustion_genes %in% rowData(cds_tcell)$gene_short_name
]
cat("Exhaustion genes available:", paste(available_genes, collapse=", "), "\n")
cat("Plotting", length(available_genes), "genes...\n")

# Feature plots on trajectory UMAP — one PNG per gene
for (gene in available_genes) {
    png(file.path(plot_dir,
                  paste0(study_id, "_trajectory_Tcell_", gene, ".png")),
        width = 7, height = 6, units = "in", res = 200, bg = "white")
    p <- plot_cells(cds_tcell,
                    genes                         = gene,
                    label_cell_groups             = FALSE,
                    label_leaves                  = FALSE,
                    label_branch_points           = FALSE,
                    cell_size                     = 0.5,
                    trajectory_graph_color        = "grey30",
                    trajectory_graph_segment_size = 0.8) +
        scale_color_viridis(option = "D", name = "Expression") +
        ggtitle(paste0(study_id, " | T Cell Trajectory — ", gene)) +
        theme_classic(base_size = 11)
    print(p)
    dev.off()
    cat(gene, "saved ✅\n")
}

cat("\nAll", length(available_genes), "gene trajectory plots saved ✅\n")
```


# STEP 13: Gene Expression Curves Along Pseudotime
```{r pseudotime_gene_curves}
# Line plots showing how exhaustion gene expression changes
# continuously along pseudotime — reveals ORDER of gene activation
# in the exhaustion program
# Expected order: TCF7 high early → TOX/PDCD1 mid → HAVCR2/CXCL13 late
# =============================================================================

library(tidyr)   # for pivot_longer

# Extract normalized expression for available exhaustion genes
expr_mat <- as.matrix(
    GetAssayData(seurat_tcell, assay = "RNA", layer = "data")[
        available_genes, , drop = FALSE
    ]
)

# Build pseudotime dataframe from T cell CDS
pt_vals <- pseudotime(cds_tcell)

# Keep only cells with finite pseudotime that exist in both objects
shared_cells <- intersect(
    names(pt_vals)[is.finite(pt_vals)],
    colnames(expr_mat)
)
cat("Shared cells for curve plotting:", length(shared_cells), "\n")

# Build combined dataframe
expr_long <- data.frame(
    t(expr_mat[, shared_cells, drop = FALSE]),
    pseudotime  = pt_vals[shared_cells],
    state       = seurat_tcell$tcell_state[shared_cells],
    hpv         = seurat_tcell$hpv[shared_cells],
    tissue      = seurat_tcell$tissue.type[shared_cells],
    stringsAsFactors = FALSE
) %>%
    pivot_longer(
        cols      = all_of(available_genes),
        names_to  = "gene",
        values_to = "expression"
    ) %>%
    mutate(gene = factor(gene, levels = available_genes))

cat("Long format rows:", nrow(expr_long), "\n")

# ------------------------------------------------------------------
# Plot 1: All exhaustion genes together — loess smoothed curves
# ------------------------------------------------------------------
p_curves <- ggplot(expr_long,
                   aes(x = pseudotime, y = expression, color = gene)) +
    geom_smooth(method  = "loess",
                span    = 0.4,
                se      = FALSE,
                linewidth = 1.1) +
    scale_color_brewer(palette = "Paired", name = "Gene") +
    labs(title    = paste0(study_id,
                           " | Exhaustion Gene Expression Along Pseudotime"),
         subtitle = "Loess-smoothed normalized expression",
         x        = "Pseudotime",
         y        = "Normalized Expression") +
    theme_classic(base_size = 12) +
    theme(plot.title    = element_text(hjust = 0.5, face = "bold"),
          plot.subtitle = element_text(hjust = 0.5, color = "grey40"),
          legend.position = "right")

png(file.path(plot_dir,
              paste0(study_id, "_trajectory_Tcell_gene_curves.png")),
    width = 12, height = 6, units = "in", res = 200, bg = "white")
print(p_curves)
dev.off()
cat("Gene curves plot saved ✅\n")

# ------------------------------------------------------------------
# Plot 2: Faceted per gene — easier to read individual dynamics
# ------------------------------------------------------------------
p_facet <- ggplot(expr_long,
                  aes(x = pseudotime, y = expression, color = gene)) +
    geom_smooth(method    = "loess",
                span      = 0.4,
                se        = TRUE,
                alpha     = 0.15,
                linewidth = 1) +
    scale_color_brewer(palette = "Paired", guide = "none") +
    facet_wrap(~gene, scales = "free_y", ncol = 5) +
    labs(title = paste0(study_id,
                        " | Exhaustion Genes Along Pseudotime (Faceted)"),
         x     = "Pseudotime",
         y     = "Normalized Expression") +
    theme_classic(base_size = 10) +
    theme(plot.title  = element_text(hjust = 0.5, face = "bold"),
          strip.text  = element_text(face = "bold"),
          axis.text.x = element_text(angle = 30, hjust = 1))

png(file.path(plot_dir,
              paste0(study_id, "_trajectory_Tcell_gene_curves_faceted.png")),
    width = 14, height = 6, units = "in", res = 200, bg = "white")
print(p_facet)
dev.off()
cat("Faceted gene curves plot saved ✅\n")

# ------------------------------------------------------------------
# Plot 3: HPV+ vs HPV– comparison — key exhaustion genes
# ------------------------------------------------------------------
hpv_genes <- c("CXCL13", "PDCD1", "HAVCR2", "TOX")
hpv_genes  <- hpv_genes[hpv_genes %in% available_genes]
cat("HPV comparison genes:", paste(hpv_genes, collapse=", "), "\n")

hpv_df <- expr_long %>%
    filter(gene %in% hpv_genes,
           !is.na(hpv),
           hpv %in% c("HPV+", "HPV-"))

if (nrow(hpv_df) > 0) {
    p_hpv_curves <- ggplot(hpv_df,
                           aes(x        = pseudotime,
                               y        = expression,
                               color    = hpv,
                               fill     = hpv,
                               linetype = hpv)) +
        geom_smooth(method    = "loess",
                    span      = 0.5,
                    se        = TRUE,
                    alpha     = 0.12,
                    linewidth = 1.1) +
        scale_color_manual(values = c("HPV+" = "#E41A1C",
                                      "HPV-" = "#377EB8"),
                           name   = "HPV Status") +
        scale_fill_manual(values  = c("HPV+" = "#E41A1C",
                                      "HPV-" = "#377EB8"),
                          guide   = "none") +
        scale_linetype_manual(values = c("HPV+" = "solid",
                                         "HPV-" = "dashed"),
                              guide  = "none") +
        facet_wrap(~gene, scales = "free_y", ncol = 2) +
        labs(title    = paste0(study_id,
                               " | Exhaustion Genes Along Pseudotime: HPV+ vs HPV–"),
             subtitle = "Loess-smoothed | shaded = 95% CI",
             x        = "Pseudotime",
             y        = "Normalized Expression") +
        theme_classic(base_size = 11) +
        theme(plot.title     = element_text(hjust = 0.5, face = "bold"),
              plot.subtitle  = element_text(hjust = 0.5, color = "grey40"),
              strip.text     = element_text(face = "bold"),
              legend.position = "top")

    png(file.path(plot_dir,
                  paste0(study_id, "_trajectory_Tcell_HPV_gene_curves.png")),
        width = 10, height = 8, units = "in", res = 200, bg = "white")
    print(p_hpv_curves)
    dev.off()
    cat("HPV gene curves plot saved ✅\n")
} else {
    cat("No HPV data found — skipping HPV curve plot\n")
}

cat("Plots saved:\n")
cat("  1.", paste0(study_id, "_trajectory_Tcell_gene_curves.png"), "\n")
cat("  2.", paste0(study_id, "_trajectory_Tcell_gene_curves_faceted.png"), "\n")
cat("  3.", paste0(study_id, "_trajectory_Tcell_HPV_gene_curves.png"), "\n")
```


# STEP 14: Reprocess CDS Through Monocle3's Own Pipeline
```{r reprocess_monocle3}
# Seurat v5 stores PCA differently — Monocle3 graph_test needs its own
# preprocess_cds → reduce_dimension pipeline to work correctly
# =============================================================================

cat("Reprocessing cds_tcell through Monocle3 pipeline...\n")

# Step 1: Preprocess within Monocle3 (PCA)
cds_tcell <- preprocess_cds(
    cds_tcell,
    num_dim   = 50,
    method    = "PCA",
    verbose   = TRUE
)
cat("Preprocessing done ✅\n")

# Step 2: Reduce dimensions within Monocle3 (UMAP)
cds_tcell <- reduce_dimension(
    cds_tcell,
    reduction_method  = "UMAP",
    preprocess_method = "PCA",
    verbose           = TRUE
)
cat("Dimension reduction done ✅\n")

# Step 3: Re-cluster cells
cds_tcell <- cluster_cells(
    cds_tcell,
    reduction_method = "UMAP",
    resolution       = 1e-3
)
cat("Clustering done ✅\n")

# Step 4: Re-learn principal graph
cat("Re-learning principal graph...\n")
cds_tcell <- learn_graph(
    cds_tcell,
    use_partition    = FALSE,
    verbose          = FALSE,
    learn_graph_control = list(
        minimal_branch_len = 5,
        ncenter           = 50
    )
)
cat("Graph learned ✅\n")

# Step 5: Re-order cells with Naive_Memory root
root_tcells <- colnames(seurat_tcell)[
    seurat_tcell$tcell_state == "Naive_Memory"
]
cat("Root cells:", length(root_tcells), "\n")

cds_tcell <- order_cells(cds_tcell, root_cells = root_tcells)

cat("Pseudotime range:",
    round(min(pseudotime(cds_tcell)[is.finite(pseudotime(cds_tcell))]), 2),
    "to",
    round(max(pseudotime(cds_tcell)[is.finite(pseudotime(cds_tcell))]), 2), "\n")

# Step 6: Fix gene_short_name again (lost after reprocessing)
rowData(cds_tcell)$gene_short_name <- rownames(cds_tcell)
cat("gene_short_name fixed ✅\n")

cat("\nReprocessing complete — now re-run Step 14 ✅\n")
```



# STEP 15: Graph-Autocorrelation Test — Genes Varying Along Pseudotime
```{r graph_test}
# Moran's I test identifies genes whose expression significantly
# changes along the trajectory (not random spatial distribution)
# High Moran's I = gene expression is structured along pseudotime
# This is the statistical backbone of trajectory DE analysis
# =============================================================================

cat("Running graph autocorrelation test (Moran's I)...\n")
cat("This may take 5-10 minutes...\n")

# Ensure gene_short_name is set (in case Step 12 was run separately)
rowData(cds_tcell)$gene_short_name <- rownames(cds_tcell)

# Run Moran's I test on T cell trajectory
cds_tcell_pr <- graph_test(
    cds_tcell,
    neighbor_graph = "principal_graph",
    cores          = 1
)

cat("Test complete ✅\n")
cat("Columns returned:", paste(names(cds_tcell_pr), collapse=", "), "\n")

# Filter significant trajectory-variable genes
traj_genes <- cds_tcell_pr %>%
    filter(q_value < 0.05, status == "OK") %>%
    arrange(desc(morans_I)) %>%
    select(gene_short_name, morans_I, q_value)

cat("\n=== Top 20 Trajectory-Variable Genes (Moran's I) ===\n")
print(head(traj_genes, 20))

# Check our key exhaustion genes specifically
cat("\n=== Exhaustion Genes in Trajectory Test ===\n")
key_genes <- c("CXCL13","PDCD1","HAVCR2","LAG3","TIGIT",
               "TCF7","TOX","FOXP3","GZMB","PRF1")
traj_genes %>%
    filter(gene_short_name %in% key_genes) %>%
    arrange(desc(morans_I)) %>%
    print()

# Save full results
write.csv(
    traj_genes,
    file.path(result_dir,
              paste0(study_id, "_trajectory_Tcell_variable_genes.csv")),
    row.names = FALSE
)

cat("\nTotal trajectory-variable genes:", nrow(traj_genes), "\n")
cat("Results saved ✅\n")
```



# STEP 16: Gene Co-expression Modules Along T Cell Trajectory
```{r gene_modules}
# Groups co-expressed trajectory-variable genes into modules.
# Each module = a gene program activated at a specific pseudotime point.
# Expected programs:
#   Module 1: Effector/Exhaustion  — CXCL13, LAG3, HAVCR2, GZMB, PRF1
#   Module 2: Proliferation        — STMN1, TYMS, TUBA1B
#   Module 3: Intermediate/Stress  — TCF7, HSP90AA1, HSPA1A
#   Module 4: Minor contamination  — FXYD3, S100A14
#   Module 5: Treg                 — FOXP3, TNFRSF18
#   Module 6: Tumor-associated     — ELF3, MAGEA4, MDK
# =============================================================================

# Fix 1: Increase resolution to get better module separation
rowData(cds_tcell)$gene_short_name <- rownames(cds_tcell)

cat("Finding gene modules with higher resolution...\n")
gene_modules <- find_gene_modules(
    cds_tcell[top_traj_genes, ],
    resolution  = 0.1,    # increased from 1e-2
    random_seed = 42
)

n_modules <- length(unique(gene_modules$module))
cat("Gene modules found:", n_modules, "\n")
cat("Module sizes:\n")
print(table(gene_modules$module))

# Exhaustion gene assignments
cat("\n=== Key Exhaustion Gene Module Assignments ===\n")
gene_modules %>%
    filter(id %in% key_genes) %>%
    arrange(module) %>%
    select(id, module) %>%
    print()

# Top 5 genes per module
cat("\n=== Top 5 Genes per Module (by Moran's I) ===\n")
gene_modules %>%
    left_join(
        traj_genes_tcell %>% rename(id = gene_short_name),
        by = "id"
    ) %>%
    group_by(module) %>%
    arrange(desc(morans_I)) %>%
    slice_head(n = 5) %>%
    select(module, id, morans_I) %>%
    print(n = Inf)

# Fix 2: aggregate_gene_expression — use named vector format for v1.4.26
# Extract cell state as named character vector
cell_state_vec <- setNames(
    as.character(seurat_tcell$tcell_state),
    colnames(seurat_tcell)
)

# Keep only cells present in cds_tcell
shared <- intersect(names(cell_state_vec), colnames(cds_tcell))
cell_state_vec <- cell_state_vec[shared]

cat("\nAggregating by T cell state...\n")
cat("Shared cells:", length(shared), "\n")

# Manual aggregation — robust alternative to aggregate_gene_expression
# Get module assignments as named vector
module_vec <- setNames(
    as.character(gene_modules$module),
    gene_modules$id
)

# Extract expression matrix
expr_cds <- as.matrix(
    exprs(cds_tcell)[gene_modules$id, shared, drop = FALSE]
)

# Aggregate: mean expression per module per cell state
states    <- unique(cell_state_vec)
state_order <- c("Naive_Memory","Progenitor_Tex","Effector","Exhausted","Treg")
states    <- state_order[state_order %in% states]
modules   <- sort(unique(gene_modules$module))

agg_mat_state <- sapply(states, function(s) {
    cells_s <- names(cell_state_vec)[cell_state_vec == s]
    cells_s <- intersect(cells_s, colnames(expr_cds))
    sapply(modules, function(m) {
        genes_m <- gene_modules$id[gene_modules$module == m]
        genes_m <- intersect(genes_m, rownames(expr_cds))
        mean(expr_cds[genes_m, cells_s, drop = FALSE], na.rm = TRUE)
    })
})
rownames(agg_mat_state) <- paste0("Module ", modules)
cat("Aggregation matrix dimensions:", dim(agg_mat_state), "\n")
print(round(agg_mat_state, 3))

# Plot heatmap — T cell states
png(file.path(plot_dir,
              paste0(study_id, "_trajectory_Tcell_modules_heatmap.png")),
    width = 9, height = 6, units = "in", res = 200, bg = "white")
pheatmap::pheatmap(
    agg_mat_state,
    scale        = "row",
    color        = colorRampPalette(
                       rev(RColorBrewer::brewer.pal(9, "RdBu"))
                   )(50),
    fontsize     = 12,
    fontsize_row = 11,
    fontsize_col = 11,
    angle_col    = 45,
    main         = paste0(study_id,
                          " | Gene Modules × T Cell Exhaustion States"),
    border_color = "white"
)
dev.off()
cat("Module × state heatmap saved ✅\n")

# Aggregate by tissue type
tissue_vec <- setNames(
    as.character(seurat_tcell$tissue.type),
    colnames(seurat_tcell)
)
tissue_vec   <- tissue_vec[shared]
tissue_order <- c("NL","LP","CA","LN")
tissues      <- tissue_order[tissue_order %in% unique(tissue_vec)]

agg_mat_tissue <- sapply(tissues, function(t) {
    cells_t <- names(tissue_vec)[tissue_vec == t]
    cells_t <- intersect(cells_t, colnames(expr_cds))
    sapply(modules, function(m) {
        genes_m <- gene_modules$id[gene_modules$module == m]
        genes_m <- intersect(genes_m, rownames(expr_cds))
        mean(expr_cds[genes_m, cells_t, drop = FALSE], na.rm = TRUE)
    })
})
rownames(agg_mat_tissue) <- paste0("Module ", modules)
cat("Tissue aggregation matrix:\n")
print(round(agg_mat_tissue, 3))

# Plot heatmap — tissue types
png(file.path(plot_dir,
              paste0(study_id, "_trajectory_Tcell_modules_tissue_heatmap.png")),
    width = 8, height = 6, units = "in", res = 200, bg = "white")
pheatmap::pheatmap(
    agg_mat_tissue,
    scale        = "row",
    color        = colorRampPalette(
                       rev(RColorBrewer::brewer.pal(9, "RdBu"))
                   )(50),
    fontsize     = 12,
    fontsize_row = 11,
    fontsize_col = 11,
    angle_col    = 45,
    main         = paste0(study_id,
                          " | Gene Modules × Tissue Type"),
    border_color = "white"
)
dev.off()
cat("Module × tissue heatmap saved ✅\n")

# Save module assignments
write.csv(
    gene_modules,
    file.path(result_dir,
              paste0(study_id, "_trajectory_Tcell_gene_modules.csv")),
    row.names = FALSE
)
cat("Module assignments saved ✅\n")
```


# STEP 17: Save Pseudotime Values + Final Summary
```{r save_pseudotime}

cat("Saving all results...\n")

# --- T cell pseudotime CSV ---
pt_tcell_out <- data.frame(
    cell_barcode = colnames(cds_tcell),
    pseudotime   = pseudotime(cds_tcell),
    state        = seurat_tcell$tcell_state,
    tissue       = seurat_tcell$tissue.type,
    hpv          = seurat_tcell$hpv,
    patient      = seurat_tcell$patient.id,
    stringsAsFactors = FALSE
)
write.csv(
    pt_tcell_out,
    file.path(result_dir,
              paste0(study_id, "_Tcell_pseudotime.csv")),
    row.names = FALSE
)
cat("T cell pseudotime CSV saved ✅\n")

# --- Full TME pseudotime CSV ---
pt_full_out <- data.frame(
    cell_barcode = colnames(cds_full),
    pseudotime   = pseudotime(cds_full),
    cell_type    = seurat_obj$cell.type,
    tissue       = seurat_obj$tissue.type,
    hpv          = seurat_obj$hpv,
    stringsAsFactors = FALSE
)
write.csv(
    pt_full_out,
    file.path(result_dir,
              paste0(study_id, "_fullTME_pseudotime.csv")),
    row.names = FALSE
)
cat("Full TME pseudotime CSV saved ✅\n")

# --- Save CDS objects ---
saveRDS(
    cds_tcell,
    file.path(result_dir,
              paste0(study_id, "_monocle3_Tcell.rds"))
)
cat("T cell CDS RDS saved ✅\n")

saveRDS(
    cds_full,
    file.path(result_dir,
              paste0(study_id, "_monocle3_fullTME.rds"))
)
cat("Full TME CDS RDS saved ✅\n")

# --- Final Summary ---
cat("\n=======================================================\n")
cat("      SCRIPT 08 — TRAJECTORY ANALYSIS SUMMARY\n")
cat("=======================================================\n")
cat("Full TME cells with finite pseudotime :",
    sum(is.finite(pseudotime(cds_full))), "\n")
cat("T cells with finite pseudotime        :",
    sum(is.finite(pseudotime(cds_tcell))), "\n")
cat("T cell trajectory-variable genes      :", nrow(traj_genes), "\n")
cat("T cell gene modules                   :",
    length(unique(gene_modules$module)), "\n")

cat("\n--- Module Interpretation ---\n")
cat("Module 1: Effector/Exhaustion program",
    "(LAG3, GZMB, PRF1, CXCL13, TIGIT, PDCD1, HAVCR2, TOX)\n")
cat("Module 2: Proliferation program",
    "(STMN1, TYMS, TUBA1B — cell cycle)\n")
cat("Module 3: Intermediate/Stress program",
    "(TCF7, HSP90AA1, HSPA1A — naive-to-exhaustion transition)\n")
cat("Module 4: Tumor-contamination signal",
    "(FXYD3, S100A14 — low expression, minor)\n")
cat("Module 5: Treg program",
    "(FOXP3, TNFRSF18, ACTB — highest in Treg)\n")
cat("Module 6: Tumor-associated signal",
    "(ELF3, MAGEA4, MDK — very low expression)\n")

cat("\n--- Disease Progression (Module 1 Exhaustion) ---\n")
cat("NL: 2.44 → LP: 4.48 → CA: 6.73 → LN: 7.42\n")
cat("3× increase in exhaustion program NL→LN\n")

cat("\nPlots saved to   :", plot_dir, "\n")
cat("Results saved to :", result_dir, "\n")
cat("\nAll outputs saved ✅\n")
cat("=======================================================\n")
```



